---
title: "Interactive Dashboard"
format:
  html:
    code-fold: true
    code-summary: "Show code"
    toc: true
    toc-depth: 3
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

# Random Walk Simulation Dashboard

This vignette demonstrates the interactive Shiny dashboard for exploring random walk simulations. The dashboard provides a user-friendly interface for configuring simulation parameters, running simulations, and visualizing results.

## Overview

The dashboard is built using Shiny modules to maintain separation between:

- **Input controls**: Parameter configuration (grid size, walkers, neighborhood type, etc.)
- **Simulation engine**: Core random walk logic (independent of GUI)
- **Output displays**: Visualization and tabulation of results

This modular architecture ensures the simulation code remains reusable outside the dashboard context.

## Features

The dashboard includes:

1. **Parameter Configuration Panel**
   - Grid size (3-100)
   - Number of walkers (dynamically limited to 60% of grid)
   - Neighborhood type (4-hood or 8-hood)
   - Boundary behavior (terminate or wrap)
   - Maximum steps per walker
   - Verbose logging option

2. **Multiple Output Views**
   - **Grid State**: Visualizes final grid with black pixels
   - **Walker Paths**: Shows trajectories and termination points
   - **Statistics**: Detailed metrics and parameter summary
   - **Raw Data**: Walker-level data and grid information

3. **Interactive Controls**
   - Run simulation with current parameters
   - Reset parameters to defaults
   - Real-time status updates
   - Progress indicators

## Architecture

### Shiny Modules

The dashboard uses three main Shiny modules:

#### 1. Input Module (`sim_input_ui` / `sim_input_server`)

Handles all parameter inputs and validation:

```{r architecture-input, eval=FALSE}
# UI creates input controls
sim_input_ui("inputs")

# Server returns reactive parameters and run trigger
inputs <- sim_input_server("inputs")
params <- inputs$params()  # Get current parameters
run_trigger <- inputs$run_trigger()  # Detect run button clicks
```

Key features:
- Dynamic validation (max walkers updates with grid size)
- Reset functionality
- Parameter bundling for simulation

#### 2. Output Module (`sim_output_ui` / `sim_output_server`)

Displays simulation results across tabs:

```{r architecture-output, eval=FALSE}
# UI creates tabbed output display
sim_output_ui("outputs")

# Server renders plots and tables from simulation result
sim_output_server("outputs", sim_result)
```

Displays:
- Grid visualization using `plot_grid()` (ggplot2)
- Walker paths using `plot_walker_paths()` (base R)
- Formatted statistics
- Data tables

#### 3. Main App (`run_dashboard`)

Orchestrates the modules:

```{r architecture-main, eval=FALSE}
run_dashboard()
```

Responsibilities:
- Combine input and output modules
- Trigger simulation on button click
- Handle errors and show notifications
- Display status messages

### Module Independence

The modules are designed to be independent:

```{r module-independence, eval=FALSE}
# Simulation code is completely independent
result <- run_simulation(
  grid_size = 20,
  n_walkers = 5,
  neighborhood = "4-hood",
  boundary = "terminate"
)

# Can be used in:
# - Command line scripts
# - Non-interactive pipelines (targets)
# - Other Shiny apps
# - Automated testing
```

## Launching the Dashboard

### Interactive Mode

To launch the dashboard interactively:

```{r launch-interactive, eval=FALSE}
library(randomwalk)

# Launch with default options
run_dashboard()

# Launch on specific port
run_dashboard(options = list(port = 3838))

# Launch and open in browser
run_dashboard(launch.browser = TRUE)
```

### Embedded in Quarto

The dashboard can also be embedded in Quarto documents using the `shiny` runtime:

````markdown
---
title: "Embedded Dashboard"
format: html
server: shiny
---

```{{r}}
library(randomwalk)
run_dashboard()
```
````

## Example Usage

### Basic Workflow

1. **Configure Parameters**
   - Adjust grid size (e.g., 30x30)
   - Set number of walkers (e.g., 10)
   - Choose neighborhood type (4-hood or 8-hood)
   - Select boundary behavior

2. **Run Simulation**
   - Click "Run Simulation"
   - Wait for completion (progress bar shows activity)
   - View results across tabs

3. **Explore Results**
   - **Grid State**: See final pattern of black pixels
   - **Walker Paths**: Examine individual walker trajectories
   - **Statistics**: Review performance metrics
   - **Raw Data**: Inspect detailed walker information

4. **Iterate**
   - Modify parameters
   - Run again to compare results
   - Reset to defaults if needed

### Parameter Effects

#### Grid Size

Larger grids allow for more complex patterns:

- Small (10x10): Quick simulations, simple patterns
- Medium (30x30): Balanced exploration
- Large (100x100): Complex fractals, longer runtime

#### Number of Walkers

More walkers create denser patterns:

- Few (1-5): Sparse, distinct paths
- Many (20+): Dense, interconnected structures

#### Neighborhood Type

Affects movement and collision detection:

- **4-hood**: Cardinal directions only (N, S, E, W)
- **8-hood**: Includes diagonals (8 possible moves)

#### Boundary Behavior

Controls edge handling:

- **Terminate**: Walkers stop at grid boundaries
- **Wrap**: Toroidal topology (edges connect)

## Testing

The dashboard modules include comprehensive tests using `shinytest2`:

```{r testing-example, eval=FALSE}
library(shinytest2)

test_that("Dashboard launches successfully", {
  app <- AppDriver$new(run_dashboard())
  expect_true(app$is_alive())
  app$stop()
})

test_that("Input module validates parameters", {
  app <- AppDriver$new(run_dashboard())

  # Set large grid
  app$set_inputs(`inputs-grid_size` = 50)

  # Check max walkers updated
  max_walkers <- floor(50 * 50 * 0.6)
  expect_equal(app$get_value(input = "inputs-n_walkers")$max, max_walkers)

  app$stop()
})
```

## Performance Considerations

### Simulation Speed

- **Grid size**: O(n²) memory, runtime varies
- **Number of walkers**: Linear impact on steps
- **Max steps**: Upper bound on computation

For large simulations (>50x50 grid, >20 walkers):

- Consider running outside dashboard
- Use verbose logging to monitor progress
- Save results for later visualization

### Dashboard Responsiveness

The dashboard uses synchronous simulation execution:

- UI blocks during simulation
- Progress bar indicates activity
- Notifications show start/completion

For async execution, see the advanced vignettes on parallel processing.

## Integration with Targets

Dashboard results can be saved and used in `targets` pipelines:

```{r targets-integration, eval=FALSE}
# _targets.R
library(targets)
library(randomwalk)

tar_plan(
  # Run simulation (outside dashboard)
  tar_target(
    sim_result,
    run_simulation(grid_size = 30, n_walkers = 10)
  ),

  # Create plots (compatible with dashboard views)
  tar_target(grid_plot, plot_grid(sim_result)),
  tar_target(paths_plot, plot_walker_paths(sim_result)),

  # Generate report including dashboard link
  tar_quarto(dashboard_doc, "inst/qmd/dashboard.qmd")
)
```

## Extending the Dashboard

### Adding New Outputs

To add a new output tab:

1. **Modify `sim_output_ui`**:
```{r extend-ui, eval=FALSE}
shiny::tabPanel(
  "New View",
  shiny::plotOutput(ns("new_plot"))
)
```

2. **Add render function in `sim_output_server`**:
```{r extend-server, eval=FALSE}
output$new_plot <- shiny::renderPlot({
  shiny::req(sim_result())
  # Custom visualization code
})
```

### Adding New Inputs

To add a new parameter:

1. **Add input control in `sim_input_ui`**:
```{r extend-input-ui, eval=FALSE}
shiny::numericInput(
  ns("new_param"),
  "New Parameter:",
  value = 10
)
```

2. **Include in params list in `sim_input_server`**:
```{r extend-input-server, eval=FALSE}
list(
  # ... existing params ...
  new_param = input$new_param
)
```

3. **Update `run_simulation` to accept new parameter**

## Summary

The Shiny dashboard provides:

- **Accessibility**: No coding required for basic use
- **Interactivity**: Real-time parameter exploration
- **Visualization**: Multiple views of results
- **Modularity**: Reusable components
- **Independence**: GUI separate from simulation logic

For programmatic use, advanced features, or batch processing, use the core `run_simulation()` function directly.

## See Also

- `?run_dashboard` - Function documentation
- `?sim_input_server` - Input module details
- `?sim_output_server` - Output module details
- `?run_simulation` - Core simulation function
- `vignette("telemetry")` - Statistical analysis
