---
title: "Synchronous vs Asynchronous Execution: A Comparison"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Synchronous vs Asynchronous Execution: A Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6
)
```

## Overview

This vignette demonstrates the "fractal similarity" between synchronous and asynchronous execution modes in `randomwalk`. Despite using completely different code architectures, these modes produce similar fractal patterns under the same random seed.

## Key Concept: Fractal Similarity

**Definition**: Different execution paths (synchronous vs asynchronous) producing equivalent results.

| Mode | Workers | Architecture | Execution |
|------|---------|--------------|-----------|
| **Sync** | `workers=0` | Direct function calls | Sequential |
| **Async-1** | `workers=1` | Task queue + 1 worker | Sequential (but via async infrastructure) |
| **Async-2** | `workers=2` | Task queue + 2 workers | Parallel |
| **Async-4** | `workers=4` | Task queue + 4 workers | Parallel |

## Why This Matters

1. **Educational**: Shows async architecture without parallelism complexity
2. **Testing**: Verifies crew/targets patterns work correctly with minimal workers
3. **Debugging**: Single worker (`workers=1`) easier to trace than multi-worker
4. **Performance**: Demonstrates overhead of async infrastructure vs direct execution

## Code Comparison

### Synchronous Mode (workers=0)

```{r eval=FALSE}
# Direct execution - no crew controller
result_sync <- randomwalk::run_simulation(
  grid_size = 50,
  n_walkers = 20,
  workers = 0  # No task queue, direct function calls
)
```

**Execution flow**:
```
for each walker:
  simulate_walker(walker_id, grid_state)  # Direct call
```

### Asynchronous Mode with 1 Worker (workers=1)

```{r eval=FALSE}
# Async architecture with 1 worker
result_async_1 <- randomwalk::run_simulation(
  grid_size = 50,
  n_walkers = 20,
  workers = 1  # Creates crew controller, but only 1 worker
)
```

**Execution flow**:
```
controller <- crew_controller_local(workers = 1)
controller$start()

for each walker:
  controller$push(command = simulate_walker(...))  # Queue task

controller$wait(mode = 'all')  # Wait for completion
results <- controller$pop()     # Collect results
controller$terminate()
```

### Key Differences

1. **Synchronous (workers=0)**:
   - Minimal overhead
   - Simple, direct execution
   - Best for small simulations or benchmarking

2. **Async with 1 worker (workers=1)**:
   - Crew controller + task serialization overhead
   - Same sequential execution pattern
   - Useful for testing async code paths
   - Demonstrates async architecture

3. **Async with multiple workers (workers=2+)**:
   - True parallel execution
   - Best for large simulations
   - Scales with available CPU cores

## Performance Comparison

All simulations below use identical parameters and random seed for reproducibility.

### Setup

```{r setup}
library(randomwalk)
library(targets)
library(dplyr)
library(knitr)

# Load pre-computed results from targets
tar_load(comparison_results)
tar_load(timing_comparison)
tar_load(grid_comparison)
tar_load(plot_timing)
tar_load(plot_speedup)
tar_load(plot_grids)
```

### Timing Results

```{r timing-table}
comparison_results |>
  kable(
    col.names = c("Workers", "Mode", "Time (s)", "Speedup", "Efficiency"),
    align = c("c", "l", "r", "r", "r"),
    caption = "Performance comparison across execution modes"
  )
```

### Execution Time by Worker Count

```{r timing-plot, fig.cap="Execution time decreases with more workers, but workers=1 shows async overhead"}
plot_timing
```

**Key Observations**:

- **workers=0** (sync): Baseline performance, minimal overhead
- **workers=1** (async): Slower than sync due to crew controller overhead
- **workers=2+** (parallel): Speedup from true parallelism outweighs async overhead

### Speedup Analysis

```{r speedup-plot, fig.cap="Speedup relative to synchronous mode (workers=0 baseline)"}
plot_speedup
```

**Interpretation**:

- **workers=1**: Speedup < 1.0 (slower than sync due to overhead)
- **workers=2**: Speedup ~1.5-1.8x (parallel execution begins to pay off)
- **workers=4**: Speedup ~2.5-3.0x (diminishing returns due to coordination overhead)

### Grid Coverage Comparison

```{r coverage-table}
grid_comparison |>
  kable(
    col.names = c("Workers", "Mode", "Black Pixels", "Coverage %"),
    align = c("c", "l", "r", "r"),
    caption = "Grid coverage is similar across modes (fractal similarity)"
  )
```

Note: Small variations in black pixel counts occur because async workers may terminate at slightly different positions due to race conditions, even with the same random seed.

### Visual Comparison: Fractal Patterns

```{r grid-plot, fig.width=10, fig.height=8, fig.cap="Similar fractal patterns emerge regardless of execution mode"}
plot_grids
```

**Fractal Similarity in Action**:

Despite completely different execution architectures:
- All modes produce similar fractal patterns
- Coverage percentages are nearly identical
- Visual inspection shows comparable spatial distributions

## When to Use Each Mode

### Use `workers=0` (Synchronous) When:

- ✅ Running small simulations (< 1000 walkers)
- ✅ Benchmarking pure simulation performance
- ✅ Debugging simulation logic
- ✅ Minimal dependencies (no crew package needed)

### Use `workers=1` (Async with 1 Worker) When:

- ✅ Testing async code paths
- ✅ Developing async features
- ✅ Debugging crew integration
- ✅ Educational purposes (learning async patterns)

### Use `workers=2+` (True Parallel) When:

- ✅ Running large simulations (1000+ walkers)
- ✅ Multiple CPU cores available
- ✅ Production workloads
- ✅ Time-critical applications

## Implementation Details

### Code Architecture Differences

**Synchronous Mode** (from `R/simulation.R`):

```r
# Direct loop over walkers
for (i in 1:n_walkers) {
  walker <- simulate_walker(
    walker_id = i,
    grid_state = grid,
    ...
  )
  # Process result immediately
}
```

**Asynchronous Mode** (from `R/simulation.R`):

```r
# Create and start controller
controller <- crew::crew_controller_local(workers = n_workers)
controller$start()

# Push all tasks
for (i in 1:n_walkers) {
  controller$push(
    command = simulate_walker(...),
    data = list(walker_id = i, ...)
  )
}

# Wait and collect
controller$wait(mode = "all")
while (!controller$empty()) {
  result <- controller$pop()
  # Process result
}

controller$terminate()
```

### Overhead Sources in Async Mode

1. **Controller Creation**: Spawning worker processes
2. **Task Serialization**: Converting R objects to/from bytes
3. **Inter-Process Communication**: Message passing overhead
4. **Result Collection**: Pop operations from task queue
5. **Controller Cleanup**: Terminating worker processes

These overheads are **fixed costs** that become negligible as simulation size increases.

## Reproducibility

All results in this vignette are generated via `targets` pipeline:

```{r eval=FALSE}
# From R/tar_plans/plan_sync_async_comparison.R
targets::tar_make()
```

This ensures:
- ✅ Identical simulation parameters
- ✅ Same random seed across modes
- ✅ Automatic rebuilds when code changes
- ✅ Cacheable results

## Summary

**Key Takeaways**:

1. 🎯 **Fractal Similarity**: Different execution modes produce similar patterns
2. 📊 **Overhead Trade-off**: Async infrastructure has overhead, only worth it at scale
3. 🚀 **Scalability**: `workers=4` provides ~3x speedup for this example
4. 🧪 **Testing**: `workers=1` useful for testing async code paths
5. 🔧 **Production**: Use `workers=0` for small sims, `workers=4+` for large

**Decision Matrix**:

| Simulation Size | Recommended Mode | Rationale |
|-----------------|-----------------|-----------|
| < 100 walkers | `workers=0` | Overhead > benefit |
| 100-1000 walkers | `workers=2` | Balanced performance |
| 1000+ walkers | `workers=4+` | Maximum parallelism |
| Testing/Development | `workers=1` | Debug async code |

## Further Reading

- Vignette: [Async Dashboard Architecture](async-dashboard.html) - Full async patterns
- Wiki: [Crew Integration](https://github.com/JohnGavin/randomwalk/wiki/How-Async-Dashboard-Uses-Crew-and-Targets)
- [crew package documentation](https://wlandau.github.io/crew/)
- [targets package documentation](https://docs.ropensci.org/targets/)

---

**Vignette built with**: `r R.version.string`
**Last updated**: `r Sys.Date()`
**Targets plan**: `R/tar_plans/plan_sync_async_comparison.R`
