---
title: "Comprehensive Sync Sequential Random Walks Dashboard"
description: "Full-featured dashboard with multiple visualizations, statistics, debug panel, and documentation links - runs in browser via WebR"
format:
  html:
    code-fold: true
    code-summary: "Show code"
    toc: false
    embed-resources: false
    include-in-header:
      - text: |
          <script>
          // Fix serviceworker_dir meta tag BEFORE Shinylive loads
          (function() {
            // Find the first meta tag (Quarto's default)
            const firstMeta = document.querySelector('meta[name="shinylive:serviceworker_dir"]');
            if (firstMeta) {
              // Update it to point to current directory
              firstMeta.setAttribute('content', '.');
            }
          })();
          </script>
    resources:
      - shinylive-sw.js
filters:
  - shinylive
---

## Dashboard Comparison

This vignette provides a **comprehensive, production-ready dashboard** with all package features. For a comparison with the simplified demo:

**This Dashboard (Comprehensive):**
- **Mode**: Both sync AND async parallel processing
- **Scope**: Full-featured with all package capabilities
- **Target**: Production-ready exploration and analysis tool
- **Controls**: Complete parameter set (workers, sync_mode, max_steps, all simulation params)
- **Visualization**: Multi-tab interface
  - Grid display with real-time updates
  - Histograms (steps, pixels, paths)
  - Individual walker paths overlay
  - Summary statistics table
  - Telemetry and performance metrics
- **Best for**: In-depth exploration, performance testing, research

**Simplified Dashboard ([dynamic_broadcasting.html](dynamic_broadcasting.html){target="_blank"}):**
- **Mode**: Synchronous only (workers=0)
- **Scope**: Single focused demonstration
- **Target**: Quick introduction to random walks
- **Controls**: Essential parameters only (grid, walkers, neighborhood, boundary)
- **Visualization**: Single grid display
- **Best for**: Learning core concepts, quick demos

---

```{shinylive-r}
#| standalone: true
#| viewerHeight: 900

# ============================================================================
# FAST Package Loading (Pre-built webR binaries from GitHub Pages)
# ============================================================================
# Packages are pre-built by GitHub Actions and served from GitHub Pages
# No compilation needed - just load directly

cat("\n=== DASHBOARD SETUP (OPTIMIZED) ===\n")
cat("Loading pre-built packages from GitHub Pages...\n\n")

# Configure repository ONCE at startup
options(repos = c(
  "https://johngavin.github.io/randomwalk",  # Our custom webR repo
  "https://repo.r-wasm.org"                   # Standard webR packages
))

# Load packages directly - NO webr::install() needed!
library(shiny)
cat("✅ shiny\n")

library(ggplot2)
cat("✅ ggplot2\n")

library(randomwalk)
cat("✅ randomwalk (v", as.character(packageVersion("randomwalk")), ")\n", sep = "")

cat("\n⚠️  WebR Mode: Synchronous only (workers=0)\n")
cat("   Reason: mirai/nanonext not available in WebAssembly\n")
cat("   Note: Async works in native R with crew backend\n")

cat("\n=== SETUP COMPLETE ===\n\n")

# ============================================================================
# Helper Functions
# ============================================================================

# WebR environment detection (from Issue #129)
is_webr <- function() {
  exists(".webr_env", envir = .GlobalEnv) ||
    identical(Sys.getenv("WEBR"), "1")
}

# ============================================================================
# Shiny UI
# ============================================================================

ui <- fluidPage(
  titlePanel("Comprehensive Async Parallel Random Walks (Issue #136)"),

  sidebarLayout(
    # LEFT PANEL: INPUTS
    sidebarPanel(
      h4("Simulation Parameters"),

      sliderInput("grid_size", "Grid Size:",
                  min = 20, max = 400, value = 100, step = 20),

      sliderInput("n_walkers", "Number of Walkers:",
                  min = 1, max = 5000, value = 2000, step = 50),

      sliderInput("workers", "Parallel Workers:",
                  min = 0, max = 4, value = 0, step = 1),
      helpText("WebR: Use 0 for synchronous mode only (async not available due to mirai/nanonext compatibility)"),

      selectInput("neighborhood", "Neighborhood:",
                  choices = c("4-hood", "8-hood"),
                  selected = "4-hood"),

      selectInput("boundary", "Boundary Behavior:",
                  choices = c("terminate", "wrap"),
                  selected = "terminate"),

      sliderInput("max_steps", "Max Steps per Walker:",
                  min = 100, max = 10000, value = 5000, step = 100),

      hr(),

      actionButton("run_sim", "Run Simulation",
                   class = "btn-primary btn-lg",
                   style = "width: 100%;"),

      hr(),

      h5("Status:"),
      verbatimTextOutput("status", placeholder = TRUE)
    ),

    # RIGHT PANEL: MULTIPLE OUTPUT PAGES
    mainPanel(
      tabsetPanel(
        id = "tabs",

        # PAGE 1: Fractal Graph (Black Pixels)
        tabPanel("Fractal Graph",
                 h4("Grid Visualization - Black Pixels"),
                 plotOutput("fractal_plot", height = "600px"),
                 hr(),
                 verbatimTextOutput("fractal_stats")),

        # PAGE 2: Walker Paths (First 25 and Last 25)
        tabPanel("Walker Paths",
                 h4("Walker Trajectories"),
                 fluidRow(
                   column(6,
                     sliderInput("first_n_paths", "First N paths:",
                                min = 0, max = 20, value = 5, step = 1)),
                   column(6,
                     sliderInput("last_n_paths", "Last N paths:",
                                min = 0, max = 20, value = 5, step = 1))
                 ),
                 helpText("Select first N and/or last N walker paths to display (max 40 total)"),
                 plotOutput("paths_plot_first25", height = "500px")),

        # PAGE 3: Distributions (Path Lengths by Termination Reason)
        tabPanel("Distributions",
                 h4("Path Length Distributions"),
                 plotOutput("dist_overall", height = "300px"),
                 h5("Overall Distribution"),
                 hr(),
                 plotOutput("dist_by_reason", height = "300px"),
                 h5("By Termination Reason")),

        # PAGE 4: Statistics
        tabPanel("Statistics",
                 h4("Comprehensive Simulation Statistics"),
                 verbatimTextOutput("stats_comprehensive")),

        # PAGE 5: Debug Info
        tabPanel("Debug",
                 h4("Debug Information"),
                 h5("Package Versions"),
                 verbatimTextOutput("debug_versions"),
                 hr(),
                 h5("Current Inputs"),
                 verbatimTextOutput("debug_inputs"),
                 hr(),
                 h5("Backend Information"),
                 verbatimTextOutput("debug_backend"),
                 hr(),
                 h5("Periodic Updates"),
                 verbatimTextOutput("debug_periodic")),

        # PAGE 6: Notes
        tabPanel("Notes",
                 h4("About This Dashboard"),
                 h5("Differences from Basic Vignette"),
                 tags$ul(
                   tags$li(strong("Comprehensive:"), " Shows all randomwalk features, not just basic demo"),
                   tags$li(strong("Multiple visualizations:"), " Fractal plots, path plots, distributions"),
                   tags$li(strong("More walkers:"), " Default 10 walkers (vs 5 in basic demo)"),
                   tags$li(strong("Detailed statistics:"), " Comprehensive stats across all tabs"),
                   tags$li(strong("Debug information:"), " Package versions, inputs, backend selection"),
                   tags$li(strong("Documentation links:"), " Cross-references to wiki and issues")
                 ),
                 hr(),
                 h5("Related Documentation"),
                 tags$ul(
                   tags$li(tags$a(href="https://github.com/JohnGavin/llm/blob/main/WIKI_CONTENT/WIKI_SHINYLIVE_LESSONS_LEARNED.md", "WIKI_SHINYLIVE_LESSONS_LEARNED.md"), " - Complete lessons learned"),
                   tags$li(tags$a(href="https://github.com/JohnGavin/randomwalk/issues/129", "Issue #129"), " - Async parallel support in WebAssembly"),
                   tags$li(tags$a(href="https://github.com/JohnGavin/randomwalk/issues/136", "Issue #136"), " - This comprehensive dashboard"),
                   tags$li(tags$a(href="https://github.com/JohnGavin/randomwalk/issues/130", "Issue #130"), " - Dynamic broadcasting (future work)"),
                   tags$li(tags$a(href="https://johngavin.github.io/randomwalk/", "randomwalk website"), " - Package documentation")
                 ),
                 hr(),
                h5("Technical Details"),
                tags$div(
                  tags$h6("WebR Environment Detection"),
                  tags$pre("is_webr <- exists('.webr_env', envir = .GlobalEnv) ||\n          identical(Sys.getenv('WEBR'), '1')"),
                  tags$h6("Backend Selection"),
                  tags$p(strong("Browser (WebR):"), " Sync sequential mode (workers=0) - one walker at a time"),
                  tags$p(strong("Native R:"), " Async parallel mode with crew backend (workers=2+)"),
                  tags$pre("if (workers > 0) {\n  if (is_webr) {\n    # mirai (not available in WebR)\n  } else {\n    # crew backend (native R)\n    controller <- crew::crew_controller_local(workers)\n  }\n}"),
                  tags$h6("Package Installation"),
                  tags$pre("webr::install('randomwalk',\n  repos = c(\n    'https://johngavin.github.io/randomwalk',\n    'https://johngavin.r-universe.dev',\n    'https://repo.r-wasm.org'\n  ))"),
                  tags$h6("Error Handling"),
                  tags$p("Comprehensive try-catch blocks with status updates and troubleshooting hints")
                ),
                 hr(),
                 h5("Known Limitations"),
                 tags$ul(
                   tags$li("No worker count UI feedback (just slider)"),
                   tags$li("No per-worker progress bars"),
                   tags$li("Dynamic broadcasting not yet implemented (Issue #130)"),
                   tags$li("First load can take 15-30 seconds (package downloads)")
                 ))
      )
    )
  )
)

# ============================================================================
# Shiny Server
# ============================================================================

server <- function(input, output, session) {

  # Reactive values for simulation state and history
  sim_result <- reactiveVal(NULL)
  sim_count <- reactiveVal(0)  # Count total simulations run
  sim_state <- reactiveVal("idle")  # States: idle, running, complete, error
  sim_start_time <- reactiveVal(NULL)
  sim_end_time <- reactiveVal(NULL)

  # Periodic timer for debug updates and status indicator
  timer <- reactiveTimer(1000)  # Update every second

  # Update status display with ticking clock during execution
  observe({
    timer()  # Depend on timer

    if (sim_state() == "running" && !is.null(sim_start_time())) {
      elapsed <- as.numeric(difftime(Sys.time(), sim_start_time(), units = "secs"))

      output$status <- renderText({
        paste0(
          "🏃 SIMULATION RUNNING...\n",
          "Run #", sim_count(), "\n",
          "━━━━━━━━━━━━━━━━━━━━━━\n",
          "Started:  ", format(sim_start_time(), "%H:%M:%S"), "\n",
          "Elapsed:  ", sprintf("%.0f", elapsed), " seconds\n",
          "Status:   ", rep(c("⏳", "⌛"), ceiling(elapsed))[elapsed %% 2 + 1], " Processing...\n",
          "━━━━━━━━━━━━━━━━━━━━━━\n",
          "Grid: ", input$grid_size, "×", input$grid_size, "\n",
          "Walkers: ", input$n_walkers, "\n",
          "Workers: ", input$workers, if (input$workers > 0) " (async)" else " (sync)"
        )
      })
    }
  })

  # Run simulation when button clicked
  observeEvent(input$run_sim, {
    # Update state to running
    sim_state("running")
    sim_start_time(Sys.time())
    sim_count(sim_count() + 1)

    output$status <- renderText({
      paste0(
        "🏃 SIMULATION RUNNING...\n",
        "Run #", sim_count(), "\n",
        "Grid: ", input$grid_size, "×", input$grid_size, "\n",
        "Walkers: ", input$n_walkers, "\n",
        "Workers: ", input$workers,
        if (input$workers > 0) " (async)" else " (sync)", "\n",
        "Started: ", format(sim_start_time(), "%H:%M:%S"), "\n",
        "Status: Processing..."
      )
    })

    tryCatch({
      start_time <- Sys.time()

      # Run simulation with async support
      result <- run_simulation(
        grid_size = input$grid_size,
        n_walkers = input$n_walkers,
        neighborhood = input$neighborhood,
        boundary = input$boundary,
        max_steps = input$max_steps,
        workers = input$workers,  # Auto-detects backend (mirai in WebR)
        sync_mode = "static",     # Grid snapshots (dynamic is Issue #130)
        verbose = FALSE,
        validate_percent = 0  # CRITICAL: Disable periodic validation (causes 100-1000x slowdown in WebR)
      )

      end_time <- Sys.time()
      elapsed <- as.numeric(difftime(end_time, start_time, units = "secs"))

      # Update state variables
      sim_state("complete")
      sim_end_time(end_time)
      sim_result(result)

      # Browser console logging for WebR diagnostics (Issue #157)
      message(sprintf("\n=== SIMULATION COMPLETE ==="))
      message(sprintf("Grid size: %dx%d", input$grid_size, input$grid_size))
      message(sprintf("Walkers: %d", input$n_walkers))
      message(sprintf("Black pixels: %d", sum(result$grid == 1)))
      message(sprintf("Elapsed time: %.2f seconds", elapsed))

      # NOTE: find_isolated_pixels() check removed due to performance impact
      # Even "once after simulation" was too slow in reactive context
      # See Issue #166 for async validation approach
      message(sprintf("==========================\n"))

      output$status <- renderText({
        paste0(
          "✅ SIMULATION COMPLETE\n",
          "Run #", sim_count(), "\n",
          "━━━━━━━━━━━━━━━━━━━━━━\n",
          "Started:  ", format(sim_start_time(), "%H:%M:%S"), "\n",
          "Finished: ", format(sim_end_time(), "%H:%M:%S"), "\n",
          "Elapsed:  ", sprintf("%.2f", elapsed), " seconds\n",
          "━━━━━━━━━━━━━━━━━━━━━━\n",
          "Backend: ", if (input$workers > 0) {
            if (is_webr()) "mirai (WebR)" else "crew (native R)"
          } else {
            "synchronous"
          }, "\n",
          "Black Pixels: ", sum(result$grid == 1), "\n",
          "Walkers: ", input$n_walkers, " completed"
        )
      })

    }, error = function(e) {
      sim_state("error")
      sim_end_time(Sys.time())

      output$status <- renderText({
        paste0(
          "❌ SIMULATION ERROR\n",
          "Run #", sim_count(), "\n",
          "━━━━━━━━━━━━━━━━━━━━━━\n",
          "Error: ", e$message, "\n",
          "━━━━━━━━━━━━━━━━━━━━━━\n",
          "Troubleshooting:\n",
          "- Check JavaScript console (F12)\n",
          "- Verify randomwalk package loaded\n",
          "- Check workers setting (0-4)\n",
          "- See WIKI_SHINYLIVE_LESSONS_LEARNED.md"
        )
      })
    })
  })

  # PAGE 1: Fractal Graph
  output$fractal_plot <- renderPlot({
    req(sim_result())
    result <- sim_result()

    # Defensive checks for WebR environment (Issue #157)
    req(!is.null(result$grid))
    req(nrow(result$grid) > 0)
    req(sum(result$grid == 1) > 0)  # Ensure has black pixels

    # CRITICAL: Validate grid connectivity before plotting
    plot_grid(result, main = "Fractal Pattern - Black Pixels on Grid")
  })

  output$fractal_stats <- renderText({
    req(sim_result())
    stats <- sim_result()$statistics
    paste0(
      "Black Pixels: ", stats$black_pixels,
      " (", sprintf("%.2f%%", stats$black_percentage), " of grid)\n",
      "Grid Size: ", stats$grid_size, "×", stats$grid_size,
      " (", stats$grid_size^2, " total cells)"
    )
  })

  # PAGE 2: Walker Paths
  output$paths_plot_first25 <- renderPlot({
    req(sim_result())
    result <- sim_result()
    n_walkers <- length(result$walkers)

    if (n_walkers >= 1) {
      first_n <- min(input$first_n_paths, n_walkers)
      last_n <- min(input$last_n_paths, n_walkers)

      # Select walker IDs: first N and last N
      walker_ids <- unique(c(
        if(first_n > 0) 1:first_n else NULL,
        if(last_n > 0) (n_walkers - last_n + 1):n_walkers else NULL
      ))

      if (length(walker_ids) > 0) {
        # Plot only selected walkers' paths
        filtered_walkers <- result$walkers[walker_ids]
        result_filtered <- result
        result_filtered$walkers <- filtered_walkers
        plot_walker_paths(result_filtered,
                          main = paste(length(walker_ids), "Selected Walker Paths"))
      } else {
        plot.new()
        text(0.5, 0.5, "No walkers selected")
      }
    } else {
      plot.new()
      text(0.5, 0.5, "No walkers available")
    }
  })

  # PAGE 3: Distributions
  output$dist_overall <- renderPlot({
    req(sim_result())
    result <- sim_result()

    # Defensive check for walkers
    if (is.null(result$walkers) || length(result$walkers) == 0) {
      plot.new()
      text(0.5, 0.5, "Run a simulation first", cex = 1.5, col = "gray")
      return()
    }

    # Extract path lengths safely
    # NOTE: walker$path is a LIST of positions, not a matrix
    # Use walker$steps for accurate step count
    path_lengths <- tryCatch({
      sapply(result$walkers, function(w) {
        if (!is.null(w$steps)) {
          w$steps
        } else {
          NA
        }
      })
    }, error = function(e) {
      numeric(0)
    })

    # Remove NAs and ensure numeric type
    path_lengths <- as.numeric(path_lengths[!is.na(path_lengths)])

    if (length(path_lengths) == 0 || !is.numeric(path_lengths)) {
      plot.new()
      text(0.5, 0.5, "No valid path data available", cex = 1.5, col = "gray")
      return()
    }

    # Need at least 2 values for meaningful distribution
    if (length(path_lengths) < 2) {
      plot.new()
      text(0.5, 0.5, paste0("Only ", length(path_lengths), " path - need at least 2"),
           cex = 1.2, col = "gray")
      return()
    }

    # Plot histogram with error handling
    tryCatch({
      hist(path_lengths,
           breaks = min(30, max(5, length(unique(path_lengths)))),
           col = "steelblue",
           border = "white",
           main = "Overall Path Length Distribution",
           xlab = "Path Length (steps)",
           ylab = "Frequency")
      abline(v = median(path_lengths), col = "red", lwd = 2, lty = 2)
      legend("topright",
             legend = c(paste("Median:", median(path_lengths)),
                       paste("Mean:", sprintf("%.1f", mean(path_lengths)))),
             col = c("red", "black"),
             lty = c(2, 1),
             lwd = 2)
    }, error = function(e) {
      plot.new()
      text(0.5, 0.5, paste0("Error creating plot:\n", e$message), cex = 1, col = "red")
    })
  })

  output$dist_by_reason <- renderPlot({
    req(sim_result())
    result <- sim_result()

    # Defensive check for walkers
    if (is.null(result$walkers) || length(result$walkers) == 0) {
      plot.new()
      text(0.5, 0.5, "Run a simulation first", cex = 1.5, col = "gray")
      return()
    }

    # Extract termination reasons and path lengths safely
    # NOTE: walker$path is a LIST, not a matrix - use walker$steps
    path_lengths <- tryCatch({
      sapply(result$walkers, function(w) {
        if (!is.null(w$steps)) w$steps else NA
      })
    }, error = function(e) numeric(0))

    reasons <- tryCatch({
      sapply(result$walkers, function(w) {
        if (!is.null(w$termination_reason)) w$termination_reason else "unknown"
      })
    }, error = function(e) character(0))

    # Remove NAs and ensure same length
    valid_idx <- !is.na(path_lengths)
    path_lengths <- as.numeric(path_lengths[valid_idx])
    reasons <- as.character(reasons[valid_idx])

    # Defensive check: ensure vectors are same length
    min_len <- min(length(path_lengths), length(reasons))
    if (min_len < length(path_lengths) || min_len < length(reasons)) {
      path_lengths <- path_lengths[1:min_len]
      reasons <- reasons[1:min_len]
    }

    # Check if we have valid data
    if (length(path_lengths) == 0 || length(reasons) == 0 ||
        !is.numeric(path_lengths) || !is.character(reasons)) {
      plot.new()
      text(0.5, 0.5, "No valid termination data available", cex = 1.5, col = "gray")
      return()
    }

    # Need at least 2 data points
    if (length(path_lengths) < 2) {
      plot.new()
      text(0.5, 0.5, "Need at least 2 walkers for distribution", cex = 1.2, col = "gray")
      return()
    }

    # Create data frame with error handling
    walkers_data <- tryCatch({
      data.frame(
        path_length = path_lengths,
        reason = reasons,
        stringsAsFactors = FALSE
      )
    }, error = function(e) {
      NULL
    })

    if (is.null(walkers_data)) {
      plot.new()
      text(0.5, 0.5, "Error creating data frame", cex = 1.2, col = "red")
      return()
    }

    # Plot boxplot by termination reason with error handling
    tryCatch({
      print(
        ggplot(walkers_data, aes(x = reason, y = path_length, fill = reason)) +
          geom_boxplot(alpha = 0.7) +
          geom_jitter(width = 0.2, alpha = 0.3) +
          labs(title = "Path Lengths by Termination Reason",
               x = "Termination Reason",
               y = "Path Length (steps)") +
          theme_minimal() +
          theme(legend.position = "none")
      )
    }, error = function(e) {
      plot.new()
      text(0.5, 0.5, paste0("Error creating plot:\n", e$message), cex = 1, col = "red")
    })
  })

  # PAGE 4: Statistics
  output$stats_comprehensive <- renderText({
    req(sim_result())
    result <- sim_result()

    # Defensive check for statistics
    if (is.null(result$statistics)) {
      return("No statistics available. Run a simulation first.")
    }

    stats <- result$statistics

    # Safely extract values with defaults - improved version
    safe_get <- function(x, default = 0) {
      tryCatch({
        if (is.null(x)) return(default)
        if (length(x) == 0) return(default)
        if (!is.atomic(x)) return(default)
        # Ensure single value for calculations
        if (length(x) > 1) return(x[1])
        return(as.numeric(x))
      }, error = function(e) {
        default
      })
    }

    # Calculate std dev safely
    std_dev <- tryCatch({
      path_lengths <- sapply(result$walkers, function(w) {
        if (!is.null(w$path) && is.matrix(w$path)) nrow(w$path) else NA
      })
      path_lengths <- as.numeric(path_lengths[!is.na(path_lengths)])
      if (length(path_lengths) > 1) {
        sd(path_lengths)
      } else {
        0
      }
    }, error = function(e) 0)

    # Extract grid size safely
    grid_size <- safe_get(stats$grid_size, default = 0)
    black_pixels <- safe_get(stats$black_pixels, default = 0)
    black_pct <- safe_get(stats$black_percentage, default = 0)
    total_walkers <- safe_get(stats$total_walkers, default = 1)
    completed_walkers <- safe_get(stats$completed_walkers, default = 0)
    total_steps <- safe_get(stats$total_steps, default = 0)
    elapsed_time <- safe_get(stats$elapsed_time_secs, default = 0.001)

    # Build output with error handling
    tryCatch({
      paste0(
        "=== GRID STATISTICS ===\n",
        "Grid Size: ", grid_size, "×", grid_size, "\n",
        "Total Cells: ", grid_size^2, "\n",
        "Black Pixels: ", black_pixels,
        sprintf(" (%.2f%%)\n", black_pct),
        "White Pixels: ", grid_size^2 - black_pixels,
        sprintf(" (%.2f%%)\n", 100 - black_pct),
        "\n=== WALKER STATISTICS ===\n",
        "Total Walkers: ", total_walkers, "\n",
        "Completed: ", completed_walkers, "\n",
        "Success Rate: ",
        sprintf("%.1f%%", 100 * completed_walkers / max(total_walkers, 1)), "\n",
        "\n=== STEP STATISTICS ===\n",
        "Total Steps: ", total_steps, "\n",
        "Min Steps: ", safe_get(stats$min_steps, 0), "\n",
        "Max Steps: ", safe_get(stats$max_steps, 0), "\n",
        "Mean Steps: ", sprintf("%.1f", safe_get(stats$mean_steps, 0)), "\n",
        "Median Steps: ", safe_get(stats$median_steps, 0), "\n",
        "Std Dev: ", sprintf("%.1f", std_dev), "\n",
        "\n=== PERFORMANCE ===\n",
        "Elapsed Time: ", sprintf("%.2f", elapsed_time), " seconds\n",
        "Steps per Second: ",
        sprintf("%.0f", total_steps / max(elapsed_time, 0.001)), "\n",
        "\n=== TERMINATION REASONS ===\n",
        if (!is.null(stats$termination_reasons) && length(stats$termination_reasons) > 0) {
          tryCatch({
            reasons_vec <- as.numeric(stats$termination_reasons)
            reasons_names <- names(stats$termination_reasons)
            if (is.null(reasons_names)) reasons_names <- paste0("Reason_", seq_along(reasons_vec))
            paste(sprintf("  %s: %d (%.1f%%)",
                          reasons_names,
                          reasons_vec,
                          100 * reasons_vec / max(total_walkers, 1)),
                  collapse = "\n")
          }, error = function(e) {
            "  Error formatting termination reasons"
          })
        } else {
          "  No data available"
        }
      )
    }, error = function(e) {
      paste0("Error displaying statistics:\n", e$message)
    })
  })

  # PAGE 5: Debug - Versions
  output$debug_versions <- renderText({
    paste0(
      "=== PACKAGE VERSIONS ===\n",
      "randomwalk: ", packageVersion("randomwalk"), "\n",
      "shiny: ", packageVersion("shiny"), "\n",
      "ggplot2: ", packageVersion("ggplot2"), "\n",
      if (requireNamespace("mirai", quietly = TRUE)) {
        paste0("mirai: ", packageVersion("mirai"), "\n")
      } else "",
      if (requireNamespace("nanonext", quietly = TRUE)) {
        paste0("nanonext: ", packageVersion("nanonext"), "\n")
      } else "",
      if (requireNamespace("crew", quietly = TRUE)) {
        paste0("crew: ", packageVersion("crew"), "\n")
      } else ""
    )
  })

  # PAGE 5: Debug - Inputs
  output$debug_inputs <- renderText({
    paste0(
      "=== CURRENT INPUTS ===\n",
      "Grid Size: ", input$grid_size, "\n",
      "Number of Walkers: ", input$n_walkers, "\n",
      "Parallel Workers: ", input$workers, "\n",
      "Neighborhood: ", input$neighborhood, "\n",
      "Boundary: ", input$boundary, "\n",
      "Max Steps: ", input$max_steps
    )
  })

  # PAGE 5: Debug - Backend
  output$debug_backend <- renderText({
    env_is_webr <- is_webr()

    paste0(
      "=== BACKEND INFORMATION ===\n",
      "Environment: ", if (env_is_webr) "WebR/Browser" else "Native R", "\n",
      "R Version: ", R.version.string, "\n",
      "Backend: ", if (input$workers > 0) {
        if (env_is_webr) "mirai (WebAssembly-compatible)" else "crew"
      } else {
        "synchronous (no parallelization)"
      }, "\n",
      "Sync Mode: static (grid snapshots)\n",
      "\n=== ENVIRONMENT DETECTION ===\n",
      ".webr_env exists: ", exists(".webr_env", envir = .GlobalEnv), "\n",
      "WEBR env var: ", Sys.getenv("WEBR"), "\n",
      "is_webr() result: ", env_is_webr
    )
  })

  # PAGE 5: Debug - Periodic Updates
  output$debug_periodic <- renderText({
    # Depend on timer to update every second
    timer()

    # Enhanced debug info during execution
    state_info <- switch(sim_state(),
      "idle" = "💤 Idle - waiting for user to click 'Run Simulation'",
      "running" = {
        elapsed <- if (!is.null(sim_start_time())) {
          as.numeric(difftime(Sys.time(), sim_start_time(), units = "secs"))
        } else {
          0
        }
        paste0("🏃 Running - ", sprintf("%.0f", elapsed), " seconds elapsed")
      },
      "complete" = {
        elapsed <- if (!is.null(sim_end_time()) && !is.null(sim_start_time())) {
          as.numeric(difftime(sim_end_time(), sim_start_time(), units = "secs"))
        } else {
          0
        }
        paste0("✅ Complete - took ", sprintf("%.2f", elapsed), " seconds")
      },
      "error" = "❌ Error - see status tab for details",
      "Unknown state"
    )

    paste0(
      "=== PERIODIC UPDATES ===\n",
      "Current Time: ", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n",
      "Simulation State: ", state_info, "\n",
      "Simulation Count: ", sim_count(), "\n",
      "Result Available: ", if (is.null(sim_result())) "No" else "Yes", "\n",
      "Active Tab: ", input$tabs, "\n",
      if (sim_state() == "running" && !is.null(sim_start_time())) {
        paste0(
          "\n=== EXECUTION INFO ===\n",
          "Grid Size: ", input$grid_size, "×", input$grid_size, "\n",
          "Walkers: ", input$n_walkers, "\n",
          "Workers: ", input$workers, "\n",
          "Backend: ", if (input$workers > 0) {
            if (exists(".webr_env", envir = .GlobalEnv) || identical(Sys.getenv("WEBR"), "1")) {
              "mirai (WebR)"
            } else {
              "crew (native R)"
            }
          } else {
            "synchronous"
          }
        )
      } else {
        ""
      }
    )
  })
}

# ============================================================================
# Run Shiny App
# ============================================================================

shinyApp(ui = ui, server = server)
```

## Comprehensive Random Walks Dashboard

This vignette provides a **comprehensive interactive Shiny application** for random walk simulations running in your browser using WebAssembly.

> **Important:** This browser-based dashboard runs in **sync sequential mode** (no parallel processing):
>
> - **WebR/Browser:** Sync sequential mode only (workers=0) - processes one walker at a time
> - **Native R (outside webpage):** The randomwalk package supports **async parallel** processing with `crew` backend (workers=2+) when run in regular R
>
> The WebR limitation is due to mirai/nanonext compatibility issues in WebAssembly. The dashboard is fully functional for visualizing random walk dynamics in the browser.

**Related Documentation:**
- [WIKI_SHINYLIVE_LESSONS_LEARNED.md](https://github.com/JohnGavin/llm/blob/main/WIKI_CONTENT/WIKI_SHINYLIVE_LESSONS_LEARNED.md) - Complete lessons learned
- [Issue #129](https://github.com/JohnGavin/randomwalk/issues/129) - Async parallel support (native R only)
- [Issue #136](https://github.com/JohnGavin/randomwalk/issues/136) - This comprehensive vignette

## Features Demonstrated

This comprehensive dashboard showcases:

### 1. **Multiple Visualization Types**
- **Fractal graphs**: Black pixel patterns on grid
- **Walker paths**: First 25 and last 25 trajectories
- **Distributions**: Overall and by termination reason
- **Statistics**: Comprehensive simulation metrics

### 2. **Async Parallel Processing**
- Auto-backend selection (mirai in WebR, crew in native R)
- True parallelization with workers=2-4
- Performance metrics and timing
- Backend information display

### 3. **Comprehensive Statistics**
- Grid statistics (black/white pixels)
- Walker statistics (completion rates)
- Step statistics (min/max/mean/median)
- Performance metrics (steps per second)
- Termination reason breakdown

### 4. **Debug Information**
- Package versions
- Current input parameters
- Backend selection details
- Environment detection results
- Periodic updates

### 5. **Documentation Integration**
- Links to wiki documentation
- Cross-references to related issues
- Implementation notes
- Known limitations

## Differences from Basic Vignette

**dynamic_broadcasting.qmd** (basic demo):
- 5 walkers (simple)
- 2 output pages (Grid, Paths, Stats)
- Minimal statistics
- Basic demo focus

**dashboard_comprehensive.qmd** (this vignette):
- 10 walkers (more comprehensive)
- 6 output pages (Fractal, Paths, Distributions, Stats, Debug, Notes)
- Comprehensive statistics and visualizations
- Production-ready showcase

## Try It Yourself

Experiment with different parameters:

- **Larger grid**: More exploration space (try 30-50)
- **More walkers**: Higher density (try 20-50)
- **More workers**: Increased parallelization (try 4)
- **8-hood**: Enable diagonal movement
- **Wrap boundary**: Toroidal grid (no boundaries)

## Related Documentation

- **Issue #129**: [Async parallel support in WebAssembly](https://github.com/JohnGavin/randomwalk/issues/129)
- **Issue #136**: [This comprehensive vignette](https://github.com/JohnGavin/randomwalk/issues/136)
- **Issue #130**: [Dynamic broadcasting via nanonext](https://github.com/JohnGavin/randomwalk/issues/130) (future work)
- **WIKI**: [WIKI_SHINYLIVE_LESSONS_LEARNED.md](https://github.com/JohnGavin/llm/blob/main/WIKI_CONTENT/WIKI_SHINYLIVE_LESSONS_LEARNED.md)
- **randomwalk Website**: https://johngavin.github.io/randomwalk/

## Session Info

```{r session-info}
sessionInfo()
```

## Version Information

```{r version-info}
#| code-fold: true
#| code-summary: "Show version details"

# Git commit information
git_commit <- tryCatch(
  system("git rev-parse HEAD", intern = TRUE),
  error = function(e) "Not available"
)
git_commit_short <- tryCatch(
  system("git rev-parse --short HEAD", intern = TRUE),
  error = function(e) "Not available"
)

# Nix environment information
nix_version <- tryCatch(
  system('nix-instantiate --eval -E "(import <nixpkgs> {}).lib.version"', intern = TRUE),
  error = function(e) "Not in Nix environment"
)

# Package version
pkg_version <- tryCatch(
  as.character(packageVersion("randomwalk")),
  error = function(e) "Package not loaded"
)

# Build timestamp
build_time <- Sys.time()

cat("=== BUILD INFORMATION ===\n\n")
cat("Git Commit:\n")
cat("  Full SHA:  ", git_commit, "\n")
cat("  Short SHA: ", git_commit_short, "\n\n")
cat("Nix Environment:\n")
cat("  nixpkgs:   ", nix_version, "\n\n")
cat("Package:\n")
cat("  randomwalk:", pkg_version, "\n")
cat("  R version: ", R.version.string, "\n\n")
cat("Build Time:  ", format(build_time, "%Y-%m-%d %H:%M:%S %Z"), "\n")
```
