---
title: "Defensive Programming in randomwalk"
author: "randomwalk package"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Show code"
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  eval = FALSE  # Examples are illustrative, not executed
)
```

# Overview

This vignette demonstrates defensive programming patterns used in the `randomwalk` package. Defensive programming helps prevent bugs, validates inputs, and provides clear error messages to users.

## Key Principles

1. **Input Validation** - Check parameters before use
2. **Graceful Degradation** - Handle errors without crashing
3. **Informative Errors** - Tell users what went wrong and how to fix it
4. **State Validation** - Verify assumptions about data structures
5. **Diagnostic Logging** - Provide debugging information when issues occur

---

# Pattern 1: Input Validation with Informative Errors

## Example: Grid Size Validation

```{r grid-validation}
#' Validate grid size parameter
#'
#' @param grid_size Integer grid dimension
#' @return Validated grid_size or error
#' @examples
#' validate_grid_size(10)   # Valid
#' validate_grid_size(-5)   # Error: must be positive
#' validate_grid_size(2.5)  # Error: must be integer
validate_grid_size <- function(grid_size) {
  # Check type
  if (!is.numeric(grid_size)) {
    stop(
      "grid_size must be numeric, got: ", class(grid_size)[1],
      call. = FALSE
    )
  }

  # Check integer
  if (grid_size != as.integer(grid_size)) {
    stop(
      "grid_size must be an integer, got: ", grid_size,
      call. = FALSE
    )
  }

  # Check positive
  if (grid_size < 1) {
    stop(
      "grid_size must be positive, got: ", grid_size,
      call. = FALSE
    )
  }

  # Check reasonable size
  if (grid_size > 1000) {
    warning(
      "Large grid_size (", grid_size, ") may cause performance issues. ",
      "Consider using grid_size <= 500 for interactive use.",
      call. = FALSE
    )
  }

  as.integer(grid_size)
}
```

**Why This Works:**
- Type check prevents crashes from wrong input types
- Range checks catch invalid values early
- Performance warnings help users avoid slow operations
- `call. = FALSE` removes confusing stack traces from errors

---

# Pattern 2: Graceful Degradation

## Example: Backend Selection with Fallback

```{r backend-fallback}
#' Select appropriate async backend for environment
#'
#' @param workers Number of parallel workers requested
#' @return Backend name ("crew", "mirai", "sync")
select_backend <- function(workers) {
  # Requested synchronous mode
  if (workers == 0) {
    return("sync")
  }

  # Check if in WebR environment
  is_webr <- function() {
    exists(".webr_env", envir = .GlobalEnv) ||
      identical(Sys.getenv("WEBR"), "1")
  }

  if (is_webr()) {
    # WebR: Try mirai, fallback to sync
    if (requireNamespace("mirai", quietly = TRUE)) {
      message("WebR detected: Using mirai backend for async processing")
      return("mirai")
    } else {
      warning(
        "Async processing requested but mirai not available in WebR. ",
        "Falling back to synchronous mode.",
        call. = FALSE
      )
      return("sync")
    }
  }

  # Native R: Use crew if available
  if (requireNamespace("crew", quietly = TRUE)) {
    message("Using crew backend for async processing")
    return("crew")
  } else {
    warning(
      "Async processing requested but crew not installed. ",
      "Install crew with: install.packages('crew'). ",
      "Falling back to synchronous mode.",
      call. = FALSE
    )
    return("sync")
  }
}
```

**Why This Works:**
- Environment detection prevents incompatibility errors
- Multiple fallback levels ensure the function always works
- Informative messages explain what's happening
- Suggests specific fixes (how to install missing packages)

---

# Pattern 3: State Validation with Diagnostics

## Example: Isolated Pixel Detection (from Issue #157)

```{r isolated-pixels}
#' Detect isolated pixels in grid that violate connectivity
#'
#' @param grid Numeric matrix
#' @param neighborhood "4-hood" or "8-hood"
#' @return Data frame of isolated pixel positions or NULL
#' @details
#' This defensive function validates grid state after async updates.
#' Isolated pixels indicate race conditions or stale state issues.
find_isolated_pixels <- function(grid, neighborhood = "4-hood") {
  if (!is.matrix(grid)) {
    stop("grid must be a matrix, got: ", class(grid)[1], call. = FALSE)
  }

  if (!is.numeric(grid)) {
    stop("grid must be numeric, got: ", typeof(grid), call. = FALSE)
  }

  rows <- nrow(grid)
  cols <- ncol(grid)
  isolated <- list()

  # Check each occupied pixel
  for (i in seq_len(rows)) {
    for (j in seq_len(cols)) {
      if (grid[i, j] != 0) {  # Occupied pixel

        # Get neighbor positions
        neighbors <- if (neighborhood == "4-hood") {
          list(
            c(i - 1, j), c(i + 1, j),
            c(i, j - 1), c(i, j + 1)
          )
        } else {  # 8-hood
          list(
            c(i - 1, j), c(i + 1, j), c(i, j - 1), c(i, j + 1),
            c(i - 1, j - 1), c(i - 1, j + 1),
            c(i + 1, j - 1), c(i + 1, j + 1)
          )
        }

        # Check if any neighbor is occupied
        has_neighbor <- FALSE
        for (nb in neighbors) {
          ni <- nb[1]
          nj <- nb[2]

          # Skip out-of-bounds neighbors
          if (ni >= 1 && ni <= rows && nj >= 1 && nj <= cols) {
            if (grid[ni, nj] != 0) {
              has_neighbor <- TRUE
              break
            }
          }
        }

        # Record isolated pixels
        if (!has_neighbor) {
          isolated[[length(isolated) + 1]] <- data.frame(
            row = i,
            col = j,
            value = grid[i, j]
          )
        }
      }
    }
  }

  if (length(isolated) == 0) {
    return(NULL)
  }

  do.call(rbind, isolated)
}
```

**Why This Works:**
- Validates input types before processing
- Checks boundary conditions (out-of-bounds neighbors)
- Returns structured diagnostic data (positions + values)
- NULL return for "no issues" allows easy conditional logic
- Clear variable names make the algorithm understandable

---

# Pattern 4: Diagnostic Logging

## Example: Async Worker Validation

```{r worker-validation}
#' Process walker results with defensive validation
#'
#' @param results List of walker results from workers
#' @param grid Current grid state
#' @param verbose Enable diagnostic logging
process_results_defensively <- function(results, grid, verbose = TRUE) {
  rejected_count <- 0
  accepted_count <- 0
  rejection_reasons <- list()

  for (result in results) {
    # Validate result structure
    if (!is.list(result) || !all(c("position", "walker_id") %in% names(result))) {
      if (verbose) {
        message("WARNING: Malformed result from worker: ",
                paste(names(result), collapse = ", "))
      }
      rejected_count <- rejected_count + 1
      rejection_reasons[[length(rejection_reasons) + 1]] <- "malformed"
      next
    }

    pos <- result$position

    # Validate position bounds
    if (pos[1] < 1 || pos[1] > nrow(grid) ||
        pos[2] < 1 || pos[2] > ncol(grid)) {
      if (verbose) {
        message("WARNING: Out-of-bounds position from walker ", result$walker_id,
                ": (", pos[1], ", ", pos[2], ")")
      }
      rejected_count <- rejected_count + 1
      rejection_reasons[[length(rejection_reasons) + 1]] <- "out_of_bounds"
      next
    }

    # Validate grid state (position should be unoccupied)
    if (grid[pos[1], pos[2]] != 0) {
      if (verbose) {
        message("DEBUG: Position (", pos[1], ", ", pos[2], ") already occupied. ",
                "Walker ", result$walker_id, " result rejected (stale state).")
      }
      rejected_count <- rejected_count + 1
      rejection_reasons[[length(rejection_reasons) + 1]] <- "occupied"
      next
    }

    # Check for isolated pixels (if neighborhood validation enabled)
    grid[pos[1], pos[2]] <- result$walker_id  # Tentatively place
    isolated <- find_isolated_pixels(grid, result$neighborhood)

    if (!is.null(isolated)) {
      if (verbose) {
        message("WARNING: Placing walker ", result$walker_id,
                " at (", pos[1], ", ", pos[2], ") would create isolated pixel(s). ",
                "Rejecting.")
      }
      grid[pos[1], pos[2]] <- 0  # Revert
      rejected_count <- rejected_count + 1
      rejection_reasons[[length(rejection_reasons) + 1]] <- "isolated_pixel"
      next
    }

    # Accept the result
    accepted_count <- accepted_count + 1
  }

  # Summary logging
  if (verbose && rejected_count > 0) {
    message("\n=== Result Validation Summary ===")
    message("Accepted: ", accepted_count)
    message("Rejected: ", rejected_count)
    message("Rejection reasons:")
    for (reason in unique(unlist(rejection_reasons))) {
      count <- sum(unlist(rejection_reasons) == reason)
      message("  - ", reason, ": ", count)
    }
  }

  list(
    grid = grid,
    accepted = accepted_count,
    rejected = rejected_count,
    rejection_reasons = table(unlist(rejection_reasons))
  )
}
```

**Why This Works:**
- Multi-level validation catches different error types
- Diagnostic logging helps debug async race conditions
- Summary statistics quantify validation effectiveness
- Verbose mode can be disabled in production
- Returns both results and diagnostic metadata

---

# Pattern 5: Assertion Helpers

## Example: Parameter Assertions

```{r assertions}
#' Assert parameter is within valid range
#'
#' @param x Parameter value
#' @param min Minimum allowed value
#' @param max Maximum allowed value
#' @param name Parameter name (for error messages)
assert_range <- function(x, min, max, name = deparse(substitute(x))) {
  if (x < min || x > max) {
    stop(
      name, " must be between ", min, " and ", max, ", got: ", x,
      call. = FALSE
    )
  }
  invisible(x)
}

#' Assert parameter is one of allowed choices
#'
#' @param x Parameter value
#' @param choices Vector of allowed values
#' @param name Parameter name
assert_choice <- function(x, choices, name = deparse(substitute(x))) {
  if (!x %in% choices) {
    stop(
      name, " must be one of: ", paste(choices, collapse = ", "),
      ", got: ", x,
      call. = FALSE
    )
  }
  invisible(x)
}

# Usage in run_simulation()
run_simulation_safe <- function(grid_size = 20,
                                n_walkers = 5,
                                neighborhood = "4-hood",
                                boundary = "terminate",
                                workers = 0) {
  # Validate all parameters upfront
  assert_range(grid_size, min = 1, max = 1000, name = "grid_size")
  assert_range(n_walkers, min = 1, max = 10000, name = "n_walkers")
  assert_range(workers, min = 0, max = 16, name = "workers")
  assert_choice(neighborhood, c("4-hood", "8-hood"), name = "neighborhood")
  assert_choice(boundary, c("terminate", "wrap"), name = "boundary")

  # Check logical constraints
  max_walkers <- floor(grid_size^2 * 0.6)
  if (n_walkers > max_walkers) {
    stop(
      "Too many walkers for grid size. Maximum ", max_walkers,
      " walkers for ", grid_size, "x", grid_size, " grid (60% of cells).",
      call. = FALSE
    )
  }

  # Proceed with validated parameters
  message("✅ All parameters validated")
  # ... rest of simulation ...
}
```

**Why This Works:**
- Reusable assertion functions reduce code duplication
- `deparse(substitute(x))` provides automatic parameter names
- `invisible(x)` allows chaining: `assert_range(assert_positive(x), 1, 100)`
- Upfront validation catches errors before expensive operations

---

# Pattern 6: Safe Defaults

## Example: Configuration with Fallbacks

```{r safe-defaults}
#' Get simulation configuration with safe defaults
#'
#' @param config User-provided configuration list
#' @return Complete configuration with validated values
get_simulation_config <- function(config = list()) {
  # Define safe defaults
  defaults <- list(
    grid_size = 20,
    n_walkers = 5,
    neighborhood = "4-hood",
    boundary = "terminate",
    workers = 0,
    max_steps = 1000,
    log_interval = 50,
    verbose = FALSE
  )

  # Merge user config with defaults
  merged <- modifyList(defaults, config)

  # Validate merged configuration
  merged$grid_size <- validate_grid_size(merged$grid_size)
  assert_range(merged$n_walkers, 1, 10000, "n_walkers")
  assert_range(merged$workers, 0, 16, "workers")
  assert_choice(merged$neighborhood, c("4-hood", "8-hood"), "neighborhood")
  assert_choice(merged$boundary, c("terminate", "wrap"), "boundary")

  # Return validated config
  merged
}

# Usage
config <- get_simulation_config(list(
  grid_size = 50,  # Override default
  workers = 4      # Override default
  # Other parameters use safe defaults
))
```

**Why This Works:**
- Users only specify what they want to change
- Missing parameters get sensible defaults
- All parameters are validated, even defaults
- Configuration is explicit and documented

---

# Best Practices Summary

## ✅ DO:

1. **Validate Early** - Check inputs at function entry
2. **Fail Fast** - Stop execution when assumptions are violated
3. **Be Specific** - Error messages should explain what's wrong and how to fix it
4. **Log Diagnostics** - Provide debugging information for complex failures
5. **Use Assertions** - Create reusable validation helpers
6. **Test Edge Cases** - Validate boundary conditions and unusual inputs
7. **Provide Defaults** - Make common use cases easy

## ❌ DON'T:

1. **Silent Failures** - Never ignore errors or return invalid data
2. **Generic Errors** - Avoid "Error: something went wrong"
3. **Assume Inputs** - Never trust user input or external data
4. **Skip Type Checks** - R's dynamic typing makes this essential
5. **Ignore Performance** - Warn users about expensive operations
6. **Crash on Warnings** - Use warnings for non-fatal issues

---

# Real-World Example: Issue #157

The WebR isolated pixels bug (Issue #157) demonstrates why defensive programming matters:

**The Problem:**
Async workers making decisions on stale grid state caused isolated pixels that violated connectivity rules.

**The Solution:**
Added defensive validation:

```{r issue-157-solution}
# Before: Trust worker results blindly
grid[pos[1], pos[2]] <- walker_id

# After: Validate before accepting
isolated <- find_isolated_pixels(grid_after_update, neighborhood)
if (!is.null(isolated)) {
  # Reject the update, log diagnostics
  message("WARNING: Update would create isolated pixels - rejecting")
  reject_update(walker_id, pos, reason = "isolated_pixel")
}
```

**Benefits:**
- Caught race conditions in async code
- Provided clear diagnostics for debugging
- Prevented data corruption
- Maintained simulation invariants

---

# Testing Defensive Code

## Unit Test Example

```{r test-validation}
test_that("validate_grid_size rejects invalid inputs", {
  # Type errors
  expect_error(
    validate_grid_size("10"),
    "grid_size must be numeric"
  )

  # Non-integer
  expect_error(
    validate_grid_size(10.5),
    "grid_size must be an integer"
  )

  # Negative
  expect_error(
    validate_grid_size(-5),
    "grid_size must be positive"
  )

  # Performance warning
  expect_warning(
    validate_grid_size(1500),
    "may cause performance issues"
  )

  # Valid input
  expect_equal(validate_grid_size(20), 20L)
})
```

---

# Conclusion

Defensive programming is essential for:

- **Reliability** - Code that handles unexpected inputs gracefully
- **Debuggability** - Clear error messages and diagnostic logging
- **Performance** - Early validation prevents expensive operations on bad data
- **User Experience** - Helpful errors guide users to solutions

The `randomwalk` package uses these patterns throughout to ensure robust async parallel processing even in challenging environments like WebR.

## Session Info

```{r session-info}
sessionInfo()
```
