---
title: "Sync Sequential Random Walks Dashboard"
format:
  html:
    code-fold: true
    code-summary: "Show code"
    toc: false
    embed-resources: false
    include-in-header:
      - text: |
          <script>
          // Fix serviceworker_dir meta tag BEFORE Shinylive loads
          (function() {
            // Find the first meta tag (Quarto's default)
            const firstMeta = document.querySelector('meta[name="shinylive:serviceworker_dir"]');
            if (firstMeta) {
              // Update it to point to current directory
              firstMeta.setAttribute('content', '.');
            }
          })();
          </script>
    resources:
      - shinylive-sw.js
filters:
  - shinylive
---

## Dashboard Comparison

This vignette provides a **simplified, focused demonstration** of synchronous random walks. For a comparison with the comprehensive dashboard:

**This Dashboard (Sync Sequential):**
- **Mode**: Synchronous only (workers=0)
- **Scope**: Single focused demo with essential controls
- **Target**: Quick demonstration of basic functionality
- **Controls**: Grid size, walkers, neighborhood, boundary
- **Visualization**: Single real-time grid display
- **Best for**: Understanding core random walk concepts

**Comprehensive Dashboard ([dashboard_comprehensive.html](dashboard_comprehensive.html){target="_blank"}):**
- **Mode**: Both sync AND async parallel modes
- **Scope**: Full-featured with all package capabilities
- **Target**: Production-ready exploration tool
- **Controls**: All parameters including worker processes, sync modes, max steps
- **Visualization**: Multiple tabs (Grid, Histograms, Paths, Statistics, Telemetry)
- **Best for**: In-depth analysis and performance comparison

---

```{shinylive-r}
#| standalone: true
#| viewerHeight: 900

# ============================================================================
# Package Installation (WebR - Synchronous Mode)
# ============================================================================
# Note: Async parallel processing disabled in WebR due to mirai/nanonext
# compatibility issues. The app runs in synchronous mode (workers=0).

cat("\n=== SYNC SEQUENTIAL SETUP (dynamic_broadcasting.qmd) ===\n")
cat("Timestamp:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n\n")

cat("Installing randomwalk...\n")
cat("Note: WebR runs in sync sequential mode (workers=0)\n\n")
tryCatch({
  webr::install(
    "randomwalk",
    repos = c(
      "https://johngavin.github.io/randomwalk",  # Fast: GitHub Pages WASM repo (5-10 min)
      "https://johngavin.r-universe.dev",        # Fallback: R-Universe (1-5 hours)
      "https://repo.r-wasm.org"                  # Standard webR packages
    )
  )
  cat("✅ randomwalk installed\n")
}, error = function(e) {
  cat("❌ ERROR installing randomwalk:", e$message, "\n")
})

cat("\nInstalling/loading required packages...\n")
# shiny is usually pre-installed in webR, but load it explicitly
library(shiny)
cat("✅ shiny loaded\n")

# Install ggplot2 dependencies and ggplot2
tryCatch({
  library(ggplot2)
  cat("✅ ggplot2 loaded\n")
}, error = function(e) {
  cat("⚠️  ggplot2 not found, installing dependencies...\n")
  # Install munsell first (required ggplot2 dependency)
  webr::install("munsell", repos = "https://repo.r-wasm.org")
  cat("✅ munsell installed\n")
  # Now install ggplot2
  webr::install("ggplot2", repos = "https://repo.r-wasm.org")
  library(ggplot2)
  cat("✅ ggplot2 installed and loaded\n")
})

# Note: Async parallel processing currently not supported in WebR
# Issue: mirai/nanonext have unresolved compatibility issues in WebAssembly
# The app will run in synchronous mode (workers=0) in WebR
cat("⚠️  Running in WebR - async parallel processing disabled\n")
cat("   Reason: mirai/nanonext compatibility issues in WebAssembly\n")
cat("   Using synchronous mode (workers=0) for browser compatibility\n")
cat("   Note: Async works in native R with crew backend\n")

tryCatch({
  library(randomwalk)
  cat("✅ randomwalk loaded\n")
  cat("Available functions:", paste(head(ls("package:randomwalk"), 10), collapse=", "), "\n")
}, error = function(e) {
  cat("❌ ERROR loading randomwalk:", e$message, "\n")
  cat("Packages available:\n")
  print(installed.packages()[, c("Package", "Version")])
})

cat("\n=== SETUP COMPLETE ===\n\n")

# ============================================================================
# Shiny UI
# ============================================================================

ui <- fluidPage(
  titlePanel("Sync Sequential Random Walks"),

  sidebarLayout(
    sidebarPanel(
      h4("Simulation Parameters"),

      sliderInput("grid_size", "Grid Size:",
                  min = 20, max = 400, value = 100, step = 20),

      sliderInput("n_walkers", "Number of Walkers:",
                  min = 1, max = 5000, value = 2000, step = 50),

      sliderInput("workers", "Parallel Workers (WebR: use 0 for sync only):",
                  min = 0, max = 4, value = 0, step = 1),

      selectInput("neighborhood", "Neighborhood:",
                  choices = c("4-hood", "8-hood"),
                  selected = "4-hood"),

      selectInput("boundary", "Boundary Behavior:",
                  choices = c("terminate", "wrap"),
                  selected = "terminate"),

      sliderInput("max_steps", "Max Steps per Walker:",
                  min = 100, max = 10000, value = 5000, step = 100),

      actionButton("run_sim", "Run Simulation",
                   class = "btn-primary btn-lg"),

      hr(),

      h5("Status:"),
      verbatimTextOutput("status")
    ),

    mainPanel(
      tabsetPanel(
        id = "tabs",

        tabPanel("Grid",
                 plotOutput("grid_plot", height = "600px")),

        tabPanel("Walker Paths",
                 fluidRow(
                   column(6,
                     sliderInput("first_n_paths", "First N paths:",
                                min = 0, max = 20, value = 5, step = 1)),
                   column(6,
                     sliderInput("last_n_paths", "Last N paths:",
                                min = 0, max = 20, value = 5, step = 1))
                 ),
                 helpText("Select first N and/or last N walker paths to display (max 40 total)"),
                 plotOutput("paths_plot", height = "600px")),

        tabPanel("Distributions",
                 h4("Path Length Distributions"),
                 plotOutput("dist_overall", height = "300px"),
                 h5("Overall Distribution"),
                 hr(),
                 plotOutput("dist_by_reason", height = "300px"),
                 h5("By Termination Reason")),

        tabPanel("Statistics",
                 verbatimTextOutput("stats_text")),

        tabPanel("Details",
                 h4("Package Information"),
                 verbatimTextOutput("package_details")),

        tabPanel("Debug",
                 verbatimTextOutput("backend_info"))
      )
    )
  )
)

# ============================================================================
# Shiny Server
# ============================================================================

server <- function(input, output, session) {

  # Reactive values for simulation state and history
  sim_result <- reactiveVal(NULL)
  sim_count <- reactiveVal(0)  # Count total simulations run
  sim_state <- reactiveVal("idle")  # States: idle, running, complete, error
  sim_start_time <- reactiveVal(NULL)
  sim_end_time <- reactiveVal(NULL)

  # Run simulation when button clicked
  observeEvent(input$run_sim, {
    # Update state to running
    sim_state("running")
    sim_start_time(Sys.time())
    sim_count(sim_count() + 1)

    output$status <- renderText({
      paste0(
        "🏃 SIMULATION RUNNING...\n",
        "Run #", sim_count(), "\n",
        "Grid: ", input$grid_size, "×", input$grid_size, "\n",
        "Walkers: ", input$n_walkers, "\n",
        "Workers: ", input$workers,
        if (input$workers > 0) " (async)" else " (sync)", "\n",
        "Started: ", format(sim_start_time(), "%H:%M:%S"), "\n",
        "Status: Processing..."
      )
    })

    tryCatch({
      start_time <- Sys.time()

      # Run simulation with async support
      result <- run_simulation(
        grid_size = input$grid_size,
        n_walkers = input$n_walkers,
        neighborhood = input$neighborhood,
        boundary = input$boundary,
        max_steps = input$max_steps,
        workers = input$workers,  # Auto-detects backend (mirai in WebR)
        sync_mode = "static",     # Grid snapshots (dynamic is Issue #130)
        verbose = FALSE
      )

      end_time <- Sys.time()
      elapsed <- as.numeric(difftime(end_time, start_time, units = "secs"))

      # Update state variables
      sim_state("complete")
      sim_end_time(end_time)
      sim_result(result)

      # Browser console logging for WebR diagnostics (Issue #157)
      message(sprintf("\n=== SIMULATION COMPLETE ==="))
      message(sprintf("Grid size: %dx%d", input$grid_size, input$grid_size))
      message(sprintf("Walkers: %d", input$n_walkers))
      message(sprintf("Black pixels: %d", sum(result$grid == 1)))
      message(sprintf("Elapsed time: %.2f seconds", elapsed))

      # NOTE: find_isolated_pixels() check removed due to performance impact
      # Even "once after simulation" was too slow in reactive context
      # See Issue #166 for async validation approach
      message(sprintf("==========================\n"))

      output$status <- renderText({
        stats <- result$statistics
        paste0(
          "✅ SIMULATION COMPLETE\n",
          "Run #", sim_count(), "\n",
          "━━━━━━━━━━━━━━━━━━━━━━\n",
          "Started:  ", format(sim_start_time(), "%H:%M:%S"), "\n",
          "Finished: ", format(sim_end_time(), "%H:%M:%S"), "\n",
          "Elapsed:  ", sprintf("%.2f", elapsed), " seconds\n",
          "━━━━━━━━━━━━━━━━━━━━━━\n",
          "Backend: ", if (input$workers > 0) "mirai" else "synchronous", "\n",
          "Black Pixels: ", stats$black_pixels,
          " (", sprintf("%.1f%%", stats$black_percentage), ")\n",
          "Walkers: ", input$n_walkers, " completed"
        )
      })

    }, error = function(e) {
      sim_state("error")
      sim_end_time(Sys.time())

      output$status <- renderText({
        paste0(
          "❌ SIMULATION ERROR\n",
          "Run #", sim_count(), "\n",
          "━━━━━━━━━━━━━━━━━━━━━━\n",
          "Error: ", e$message, "\n",
          "━━━━━━━━━━━━━━━━━━━━━━\n",
          "Check JavaScript console (F12) for details"
        )
      })
    })
  })

  # Grid plot
  output$grid_plot <- renderPlot({
    req(sim_result())
    result <- sim_result()

    # Defensive checks for WebR environment (Issue #157)
    req(!is.null(result$grid))
    req(nrow(result$grid) > 0)
    req(sum(result$grid == 1) > 0)  # Ensure has black pixels

    plot_grid(result, main = "Final Grid State")
  })

  # Walker paths plot
  output$paths_plot <- renderPlot({
    req(sim_result())
    result <- sim_result()
    n_walkers <- length(result$walkers)

    if (n_walkers >= 1) {
      first_n <- min(input$first_n_paths, n_walkers)
      last_n <- min(input$last_n_paths, n_walkers)

      # Select walker IDs: first N and last N
      walker_ids <- unique(c(
        if(first_n > 0) 1:first_n else NULL,
        if(last_n > 0) (n_walkers - last_n + 1):n_walkers else NULL
      ))

      if (length(walker_ids) > 0) {
        # Plot only selected walkers' paths
        filtered_walkers <- result$walkers[walker_ids]
        result_filtered <- result
        result_filtered$walkers <- filtered_walkers
        plot_walker_paths(result_filtered)
      } else {
        plot.new()
        text(0.5, 0.5, "No walkers selected")
      }
    } else {
      plot.new()
      text(0.5, 0.5, "No walkers available")
    }
  })

  # Distributions
  output$dist_overall <- renderPlot({
    req(sim_result())
    result <- sim_result()

    # Defensive check for walkers
    if (is.null(result$walkers) || length(result$walkers) == 0) {
      plot.new()
      text(0.5, 0.5, "Run a simulation first", cex = 1.5, col = "gray")
      return()
    }

    # Extract path lengths safely
    # NOTE: walker$path is a LIST, not a matrix - use walker$steps
    path_lengths <- tryCatch({
      sapply(result$walkers, function(w) {
        if (!is.null(w$steps)) w$steps else NA
      })
    }, error = function(e) numeric(0))

    path_lengths <- path_lengths[!is.na(path_lengths)]

    if (length(path_lengths) == 0) {
      plot.new()
      text(0.5, 0.5, "No path data available", cex = 1.5, col = "gray")
      return()
    }

    # Plot histogram
    hist(path_lengths,
         breaks = 30,
         col = "steelblue",
         border = "white",
         main = "Overall Path Length Distribution",
         xlab = "Path Length (steps)",
         ylab = "Frequency")
    abline(v = median(path_lengths), col = "red", lwd = 2, lty = 2)
    legend("topright",
           legend = c(paste("Median:", median(path_lengths)),
                     paste("Mean:", sprintf("%.1f", mean(path_lengths)))),
           col = c("red", "black"),
           lty = c(2, 1),
           lwd = 2)
  })

  output$dist_by_reason <- renderPlot({
    req(sim_result())
    result <- sim_result()

    # Defensive check for walkers
    if (is.null(result$walkers) || length(result$walkers) == 0) {
      plot.new()
      text(0.5, 0.5, "Run a simulation first", cex = 1.5, col = "gray")
      return()
    }

    # Extract data safely
    # NOTE: walker$path is a LIST, not a matrix - use walker$steps
    path_lengths <- tryCatch({
      sapply(result$walkers, function(w) {
        if (!is.null(w$steps)) w$steps else NA
      })
    }, error = function(e) numeric(0))

    reasons <- tryCatch({
      sapply(result$walkers, function(w) {
        if (!is.null(w$termination_reason)) w$termination_reason else "unknown"
      })
    }, error = function(e) character(0))

    # Remove NAs
    valid_idx <- !is.na(path_lengths)
    path_lengths <- path_lengths[valid_idx]
    reasons <- reasons[valid_idx]

    if (length(path_lengths) == 0 || length(reasons) == 0) {
      plot.new()
      text(0.5, 0.5, "No termination data available", cex = 1.5, col = "gray")
      return()
    }

    walkers_data <- data.frame(
      path_length = path_lengths,
      reason = reasons,
      stringsAsFactors = FALSE
    )

    # Plot boxplot by termination reason
    ggplot(walkers_data, aes(x = reason, y = path_length, fill = reason)) +
      geom_boxplot(alpha = 0.7) +
      geom_jitter(width = 0.2, alpha = 0.3) +
      labs(title = "Path Lengths by Termination Reason",
           x = "Termination Reason",
           y = "Path Length (steps)") +
      theme_minimal() +
      theme(legend.position = "none")
  })

  # Statistics text
  output$stats_text <- renderText({
    req(sim_result())
    stats <- sim_result()$statistics

    paste0(
      "Grid Size: ", stats$grid_size, "×", stats$grid_size, "\n",
      "Black Pixels: ", stats$black_pixels,
      sprintf(" (%.2f%%)\n", stats$black_percentage),
      "\nWalkers:\n",
      "  Total: ", stats$total_walkers, "\n",
      "  Completed: ", stats$completed_walkers, "\n",
      "\nSteps:\n",
      "  Total: ", stats$total_steps, "\n",
      "  Min: ", stats$min_steps, "\n",
      "  Max: ", stats$max_steps, "\n",
      "  Mean: ", sprintf("%.1f", stats$mean_steps), "\n",
      "  Median: ", stats$median_steps, "\n",
      "\nPerformance:\n",
      "  Elapsed Time: ", sprintf("%.2f", stats$elapsed_time_secs), "s\n",
      "\nTermination Reasons:\n",
      paste(sprintf("  %s: %d", names(stats$termination_reasons),
                    stats$termination_reasons), collapse = "\n")
    )
  })

  # Backend information with dynamic simulation data
  output$backend_info <- renderText({
    is_webr <- exists(".webr_env", envir = .GlobalEnv) ||
               identical(Sys.getenv("WEBR"), "1")

    # Static environment info
    env_info <- paste0(
      "=== ENVIRONMENT ===\n",
      "Runtime: ", if (is_webr) "WebR/Browser" else "Native R", "\n",
      "Mode: ", if (input$workers > 0) {
        if (is_webr) "mirai (not available - running sync sequential)" else "async parallel (crew)"
      } else {
        if (is_webr) "sync sequential (browser limitation)" else "sync sequential (workers=0)"
      }, "\n",
      "\n=== PACKAGE VERSIONS ===\n",
      "randomwalk: ", packageVersion("randomwalk"), "\n",
      if (requireNamespace("mirai", quietly = TRUE)) {
        paste0("mirai: ", packageVersion("mirai"), "\n")
      } else "",
      if (requireNamespace("nanonext", quietly = TRUE)) {
        paste0("nanonext: ", packageVersion("nanonext"), "\n")
      } else "",
      if (requireNamespace("crew", quietly = TRUE)) {
        paste0("crew: ", packageVersion("crew"), "\n")
      } else ""
    )

    # Dynamic simulation data (if available)
    result <- sim_result()
    if (is.null(result)) {
      paste0(
        env_info,
        "\n=== SIMULATION DATA ===\n",
        "No simulation run yet - click 'Run Simulation' to generate data\n",
        "\n=== STATUS ===\n",
        "State: ", sim_state(), "\n",
        "Total Runs: ", sim_count()
      )
    } else {
      stats <- result$statistics

      paste0(
        env_info,
        "\n=== SIMULATION DATA (Run #", sim_count(), ") ===\n",
        "State: ", sim_state(), "\n",
        if (!is.null(sim_start_time()) && !is.null(sim_end_time())) {
          paste0(
            "Started: ", format(sim_start_time(), "%H:%M:%S"), "\n",
            "Finished: ", format(sim_end_time(), "%H:%M:%S"), "\n",
            "Elapsed: ", sprintf("%.2f", as.numeric(difftime(sim_end_time(), sim_start_time(), units = "secs"))), " seconds\n"
          )
        } else "",
        "\n=== GRID STATE ===\n",
        "Size: ", stats$grid_size, "×", stats$grid_size, "\n",
        "Black Pixels: ", stats$black_pixels, " (", sprintf("%.1f%%", stats$black_percentage), ")\n",
        "White Pixels: ", stats$grid_size^2 - stats$black_pixels, " (", sprintf("%.1f%%", 100 - stats$black_percentage), ")\n",
        "\n=== WALKER STATISTICS ===\n",
        "Total: ", stats$total_walkers, "\n",
        "Completed: ", stats$completed_walkers, "\n",
        "Steps (Total): ", stats$total_steps, "\n",
        "Steps (Mean): ", sprintf("%.1f", stats$mean_steps), "\n",
        "Steps (Median): ", stats$median_steps, "\n",
        "Steps (Min): ", stats$min_steps, "\n",
        "Steps (Max): ", stats$max_steps, "\n",
        "\n=== TERMINATION REASONS ===\n",
        if (!is.null(stats$termination_reasons) && length(stats$termination_reasons) > 0) {
          paste(sprintf("%s: %d (%.1f%%)",
                        names(stats$termination_reasons),
                        stats$termination_reasons,
                        100 * stats$termination_reasons / stats$total_walkers),
                collapse = "\n")
        } else {
          "No data"
        },
        "\n=== PERFORMANCE ===\n",
        "Elapsed: ", sprintf("%.2f", stats$elapsed_time_secs), " seconds\n",
        "Steps/Second: ", sprintf("%.0f", stats$total_steps / max(stats$elapsed_time_secs, 0.001)), "\n"
      )
    }
  })

  # Package details
  output$package_details <- renderText({
    paste0(
      "=== PACKAGE SOURCES ===\n\n",
      "randomwalk:\n",
      "  https://johngavin.github.io/randomwalk/\n\n",
      "WebR packages (NOT installed - parallel processing disabled):\n",
      "  nanonext & mirai: https://r-lib.r-universe.dev\n",
      "  Note: Not compatible with WebR - causes browser version to run sync sequential\n\n",
      "=== PROCESSING MODES ===\n\n",
      "THIS BROWSER VERSION (WebR):\n",
      "  - Sync Sequential Mode (workers=0)\n",
      "  - Processes one walker at a time sequentially\n",
      "  - No parallel processing available\n",
      "  - All computation happens client-side in browser\n\n",
      "R CODE OUTSIDE WEBPAGE (Native R):\n",
      "  - Async Parallel Mode (workers=2+)\n",
      "  - Uses crew backend for async parallelization\n",
      "  - crew internally uses mirai for parallel processing\n",
      "  - Can process multiple walkers simultaneously\n",
      "  - Full async/parallel capabilities when running randomwalk package\n",
      "    in regular R (not in browser)\n\n",
      "=== SYNC MODES (Both Environments) ===\n\n",
      "sync_mode='static' (current):\n",
      "  - Workers get grid snapshots\n",
      "  - Used in native R async parallel mode\n",
      "  - Not applicable in browser (no workers)\n\n",
      "sync_mode='dynamic' (future):\n",
      "  - Real-time broadcasting via nanonext\n",
      "  - Issue #130 - future work\n",
      "  - Would require nanonext WebR compatibility"
    )
  })
}

# ============================================================================
# Run Shiny App
# ============================================================================

shinyApp(ui = ui, server = server)
```

## Sync Sequential Random Walks

This vignette provides a **fully interactive Shiny application** for sync sequential random walk simulations running directly in your browser using WebAssembly.

> **Important:** This browser-based dashboard runs in **sync sequential mode** (no parallel processing):
>
> - **WebR/Browser:** Sync sequential mode only (workers=0) - processes one walker at a time
> - **Native R (outside webpage):** The randomwalk package supports **async parallel** processing with `crew` backend (workers=2+) when run in regular R
>
> The WebR limitation is due to mirai/nanonext compatibility issues in WebAssembly (tracked in Issue #158). The dashboard is fully functional for visualizing random walk dynamics in the browser.

## Features

This dashboard demonstrates:

1. **Browser-based Simulation**: Runs entirely in your browser via WebAssembly (no server required)
2. **Real-time Visualization**: Grid state, walker paths, and statistics
3. **Parameter Tuning**: Adjust grid size, walkers, boundaries interactively
4. **Path Filtering**: View first N and/or last N walker paths (Details page for technical info)

## Notes

- **This Browser Version**: Sync sequential mode (workers=0) - processes walkers one at a time
- **R Code Outside Webpage**: Full async parallel support with crew backend (workers=2+) when running randomwalk package in native R
- See **Details** tab in the app for technical information about implementation differences

## Try It Yourself

Experiment with different parameters:

- **Larger grid**: More exploration space (try 30 or 40)
- **More walkers**: Increase density (try 10 or 15)
- **Path filters**: Adjust first_n/last_n sliders on Walker Paths tab
- **8-hood**: Enable diagonal movement for more path options
- **Wrap boundary**: Toroidal grid (no boundaries)

Note: Keep workers=0 in WebR mode

## Session Info

```{r session-info}
sessionInfo()
```

## Version Information

```{r version-info}
#| code-fold: true
#| code-summary: "Show version details"

# Git commit information
git_commit <- tryCatch(
  system("git rev-parse HEAD", intern = TRUE),
  error = function(e) "Not available"
)
git_commit_short <- tryCatch(
  system("git rev-parse --short HEAD", intern = TRUE),
  error = function(e) "Not available"
)

# Nix environment information
nix_version <- tryCatch(
  system('nix-instantiate --eval -E "(import <nixpkgs> {}).lib.version"', intern = TRUE),
  error = function(e) "Not in Nix environment"
)

# Package version
pkg_version <- tryCatch(
  as.character(packageVersion("randomwalk")),
  error = function(e) "Package not loaded"
)

# Build timestamp
build_time <- Sys.time()

cat("=== BUILD INFORMATION ===\n\n")
cat("Git Commit:\n")
cat("  Full SHA:  ", git_commit, "\n")
cat("  Short SHA: ", git_commit_short, "\n\n")
cat("Nix Environment:\n")
cat("  nixpkgs:   ", nix_version, "\n\n")
cat("Package:\n")
cat("  randomwalk:", pkg_version, "\n")
cat("  R version: ", R.version.string, "\n\n")
cat("Build Time:  ", format(build_time, "%Y-%m-%d %H:%M:%S %Z"), "\n")
```
