---
title: "Sync Sequential Random Walks Dashboard"
description: "Interactive Shiny dashboard for sequential random walk simulations with real-time visualization - runs in browser via WebR"
format:
  html:
    code-fold: true
    code-summary: "Show code"
    toc: false
    embed-resources: false
    include-in-header:
      - text: |
          <script>
          // Fix serviceworker_dir meta tag BEFORE Shinylive loads
          (function() {
            // Find the first meta tag (Quarto's default)
            const firstMeta = document.querySelector('meta[name="shinylive:serviceworker_dir"]');
            if (firstMeta) {
              // Update it to point to current directory
              firstMeta.setAttribute('content', '.');
            }
          })();
          </script>
    resources:
      - shinylive-sw.js
filters:
  - shinylive
---

```{shinylive-r}
#| standalone: true
#| viewerHeight: 900

# ============================================================================
# FAST Package Loading (Pre-built webR binaries from GitHub Pages)
# ============================================================================
# Packages built by GitHub Actions and deployed to GitHub Pages
# See: https://github.com/JohnGavin/randomwalk/wiki/Shinylive-Lessons-Learned

cat("\n=== DASHBOARD SETUP ===\n")
cat("Installing packages from GitHub Pages...\n\n")

# Install randomwalk from GitHub Pages (built by deploy-pages.yaml workflow)
tryCatch({
  webr::install(
    "randomwalk",
    repos = c(
      "https://johngavin.github.io/randomwalk",  # GitHub Pages webR repo
      "https://repo.r-wasm.org"                   # Standard webR packages
    )
  )
  cat("✅ randomwalk installed\n")
}, error = function(e) {
  cat("❌ ERROR installing randomwalk:", e$message, "\n")
})

# Load shiny (usually pre-installed)
library(shiny)
cat("✅ shiny\n")

# Install ggplot2 dependencies and ggplot2
tryCatch({
  library(ggplot2)
  cat("✅ ggplot2\n")
}, error = function(e) {
  cat("⚠️  ggplot2 not found, installing dependencies...\n")
  # Install munsell first (required ggplot2 dependency often missing)
  webr::install("munsell", repos = "https://repo.r-wasm.org")
  cat("✅ munsell installed\n")
  webr::install("ggplot2", repos = "https://repo.r-wasm.org")
  library(ggplot2)
  cat("✅ ggplot2 installed and loaded\n")
})

# Load randomwalk
tryCatch({
  library(randomwalk)
  cat("✅ randomwalk (v", as.character(packageVersion("randomwalk")), ")\n", sep = "")
}, error = function(e) {
  cat("❌ ERROR loading randomwalk:", e$message, "\n")
  cat("Available packages:\n")
  print(installed.packages()[, c("Package", "Version")])
})

cat("\n⚠️  WebR Mode: Synchronous only (workers=0)\n")
cat("   Reason: mirai/nanonext not available in WebAssembly\n")
cat("   Note: Async works in native R with crew backend\n")

cat("\n=== SETUP COMPLETE ===\n\n")

# ============================================================================
# Shiny UI
# ============================================================================

ui <- fluidPage(
  titlePanel("Sync Sequential Random Walks"),

  sidebarLayout(
    sidebarPanel(
      h4("Simulation Parameters"),

      sliderInput("grid_size", "Grid Size:",
                  min = 20, max = 400, value = 100, step = 20),

      sliderInput("n_walkers", "Number of Walkers:",
                  min = 1, max = 5000, value = 2000, step = 50),

      selectInput("neighborhood", "Neighborhood:",
                  choices = c("4-hood", "8-hood"),
                  selected = "4-hood"),

      selectInput("boundary", "Boundary Behavior:",
                  choices = c("terminate", "wrap"),
                  selected = "terminate"),

      sliderInput("max_steps", "Max Steps per Walker:",
                  min = 100, max = 10000, value = 5000, step = 100),

      actionButton("run_sim", "Run Simulation",
                   class = "btn-primary btn-lg"),

      hr(),

      h5("Status:"),
      verbatimTextOutput("status")
    ),

    mainPanel(
      tabsetPanel(
        id = "tabs",

        tabPanel("Grid",
                 plotOutput("grid_plot", height = "600px")),

        tabPanel("Walker Paths",
                 fluidRow(
                   column(6,
                     sliderInput("first_n_paths", "First N paths:",
                                min = 0, max = 20, value = 5, step = 1)),
                   column(6,
                     sliderInput("last_n_paths", "Last N paths:",
                                min = 0, max = 20, value = 5, step = 1))
                 ),
                 helpText("Select first N and/or last N walker paths to display (max 40 total)"),
                 plotOutput("paths_plot", height = "600px")),

        tabPanel("Distributions",
                 h4("Path Length Distributions"),
                 plotOutput("dist_overall", height = "300px"),
                 h5("Overall Distribution"),
                 hr(),
                 plotOutput("dist_by_reason", height = "300px"),
                 h5("By Termination Reason")),

        tabPanel("Statistics",
                 verbatimTextOutput("stats_text")),

        tabPanel("Details",
                 h4("Package Information"),
                 verbatimTextOutput("package_details")),

        tabPanel("Debug",
                 verbatimTextOutput("backend_info"))
      )
    )
  )
)

# ============================================================================
# Shiny Server
# ============================================================================

server <- function(input, output, session) {

  # Reactive values for simulation state and history
  sim_result <- reactiveVal(NULL)
  sim_count <- reactiveVal(0)  # Count total simulations run
  sim_state <- reactiveVal("idle")  # States: idle, running, complete, error
  sim_start_time <- reactiveVal(NULL)
  sim_end_time <- reactiveVal(NULL)

  # Run simulation when button clicked
  observeEvent(input$run_sim, {
    # Update state to running
    sim_state("running")
    sim_start_time(Sys.time())
    sim_count(sim_count() + 1)

    output$status <- renderText({
      paste0(
        "🏃 SIMULATION RUNNING...\n",
        "Run #", sim_count(), "\n",
        "Grid: ", input$grid_size, "×", input$grid_size, "\n",
        "Walkers: ", input$n_walkers, "\n",
        "Workers: 0 (sync sequential)\n",
        "Started: ", format(sim_start_time(), "%H:%M:%S"), "\n",
        "Status: Processing..."
      )
    })

    tryCatch({
      start_time <- Sys.time()

      # Run simulation with async support
      result <- run_simulation(
        grid_size = input$grid_size,
        n_walkers = input$n_walkers,
        neighborhood = input$neighborhood,
        boundary = input$boundary,
        max_steps = input$max_steps,
        workers = 0,  # WebR: sync sequential only (no parallel in browser)
        sync_mode = "static",
        verbose = FALSE,
        validate_percent = 0  # CRITICAL: Disable periodic validation (causes 100-1000x slowdown in WebR)
      )

      end_time <- Sys.time()
      elapsed <- as.numeric(difftime(end_time, start_time, units = "secs"))

      # Update state variables
      sim_state("complete")
      sim_end_time(end_time)
      sim_result(result)

      # Browser console logging for WebR diagnostics (Issue #157)
      message(sprintf("\n=== SIMULATION COMPLETE ==="))
      message(sprintf("Grid size: %dx%d", input$grid_size, input$grid_size))
      message(sprintf("Walkers: %d", input$n_walkers))
      message(sprintf("Black pixels: %d", sum(result$grid == 1)))
      message(sprintf("Elapsed time: %.2f seconds", elapsed))

      # NOTE: find_isolated_pixels() check removed due to performance impact
      # Even "once after simulation" was too slow in reactive context
      # See Issue #166 for async validation approach
      message(sprintf("==========================\n"))

      output$status <- renderText({
        stats <- result$statistics
        paste0(
          "✅ SIMULATION COMPLETE\n",
          "Run #", sim_count(), "\n",
          "━━━━━━━━━━━━━━━━━━━━━━\n",
          "Started:  ", format(sim_start_time(), "%H:%M:%S"), "\n",
          "Finished: ", format(sim_end_time(), "%H:%M:%S"), "\n",
          "Elapsed:  ", sprintf("%.2f", elapsed), " seconds\n",
          "━━━━━━━━━━━━━━━━━━━━━━\n",
          "Backend: synchronous (WebR browser)\n",
          "Black Pixels: ", stats$black_pixels,
          " (", sprintf("%.1f%%", stats$black_percentage), ")\n",
          "Walkers: ", input$n_walkers, " completed"
        )
      })

    }, error = function(e) {
      sim_state("error")
      sim_end_time(Sys.time())

      output$status <- renderText({
        paste0(
          "❌ SIMULATION ERROR\n",
          "Run #", sim_count(), "\n",
          "━━━━━━━━━━━━━━━━━━━━━━\n",
          "Error: ", e$message, "\n",
          "━━━━━━━━━━━━━━━━━━━━━━\n",
          "Check JavaScript console (F12) for details"
        )
      })
    })
  })

  # Grid plot
  output$grid_plot <- renderPlot({
    req(sim_result())
    result <- sim_result()

    # Defensive checks for WebR environment (Issue #157)
    req(!is.null(result$grid))
    req(nrow(result$grid) > 0)
    req(sum(result$grid == 1) > 0)  # Ensure has black pixels

    plot_grid(result, main = "Final Grid State", show_legend = FALSE)
  })

  # Walker paths plot
  output$paths_plot <- renderPlot({
    req(sim_result())
    result <- sim_result()
    n_walkers <- length(result$walkers)

    if (n_walkers >= 1) {
      first_n <- min(input$first_n_paths, n_walkers)
      last_n <- min(input$last_n_paths, n_walkers)

      # Select walker IDs: first N and last N
      walker_ids <- unique(c(
        if(first_n > 0) 1:first_n else NULL,
        if(last_n > 0) (n_walkers - last_n + 1):n_walkers else NULL
      ))

      if (length(walker_ids) > 0) {
        # Plot only selected walkers' paths
        filtered_walkers <- result$walkers[walker_ids]
        result_filtered <- result
        result_filtered$walkers <- filtered_walkers
        plot_walker_paths(result_filtered)
      } else {
        plot.new()
        text(0.5, 0.5, "No walkers selected")
      }
    } else {
      plot.new()
      text(0.5, 0.5, "No walkers available")
    }
  })

  # Distributions
  output$dist_overall <- renderPlot({
    req(sim_result())
    result <- sim_result()

    # Defensive check for walkers
    if (is.null(result$walkers) || length(result$walkers) == 0) {
      plot.new()
      text(0.5, 0.5, "Run a simulation first", cex = 1.5, col = "gray")
      return()
    }

    # Extract path lengths safely
    # NOTE: walker$path is a LIST, not a matrix - use walker$steps
    path_lengths <- tryCatch({
      sapply(result$walkers, function(w) {
        if (!is.null(w$steps)) w$steps else NA
      })
    }, error = function(e) numeric(0))

    path_lengths <- path_lengths[!is.na(path_lengths)]

    if (length(path_lengths) == 0) {
      plot.new()
      text(0.5, 0.5, "No path data available", cex = 1.5, col = "gray")
      return()
    }

    # Plot histogram
    hist(path_lengths,
         breaks = 30,
         col = "steelblue",
         border = "white",
         main = "Overall Path Length Distribution",
         xlab = "Path Length (steps)",
         ylab = "Frequency")
    abline(v = median(path_lengths), col = "red", lwd = 2, lty = 2)
    legend("topright",
           legend = c(paste("Median:", median(path_lengths)),
                     paste("Mean:", sprintf("%.1f", mean(path_lengths)))),
           col = c("red", "black"),
           lty = c(2, 1),
           lwd = 2)
  })

  output$dist_by_reason <- renderPlot({
    req(sim_result())
    result <- sim_result()

    # Defensive check for walkers
    if (is.null(result$walkers) || length(result$walkers) == 0) {
      plot.new()
      text(0.5, 0.5, "Run a simulation first", cex = 1.5, col = "gray")
      return()
    }

    # Extract data safely
    # NOTE: walker$path is a LIST, not a matrix - use walker$steps
    path_lengths <- tryCatch({
      sapply(result$walkers, function(w) {
        if (!is.null(w$steps)) w$steps else NA
      })
    }, error = function(e) numeric(0))

    reasons <- tryCatch({
      sapply(result$walkers, function(w) {
        if (!is.null(w$termination_reason)) w$termination_reason else "unknown"
      })
    }, error = function(e) character(0))

    # Remove NAs
    valid_idx <- !is.na(path_lengths)
    path_lengths <- path_lengths[valid_idx]
    reasons <- reasons[valid_idx]

    if (length(path_lengths) == 0 || length(reasons) == 0) {
      plot.new()
      text(0.5, 0.5, "No termination data available", cex = 1.5, col = "gray")
      return()
    }

    walkers_data <- data.frame(
      path_length = path_lengths,
      reason = reasons,
      stringsAsFactors = FALSE
    )

    # Plot boxplot by termination reason
    ggplot(walkers_data, aes(x = reason, y = path_length, fill = reason)) +
      geom_boxplot(alpha = 0.7) +
      geom_jitter(width = 0.2, alpha = 0.3) +
      labs(title = "Path Lengths by Termination Reason",
           x = "Termination Reason",
           y = "Path Length (steps)") +
      theme_minimal() +
      theme(legend.position = "none")
  })

  # Statistics text
  output$stats_text <- renderText({
    req(sim_result())
    stats <- sim_result()$statistics

    paste0(
      "Grid Size: ", stats$grid_size, "×", stats$grid_size, "\n",
      "Black Pixels: ", stats$black_pixels,
      sprintf(" (%.2f%%)\n", stats$black_percentage),
      "\nWalkers:\n",
      "  Total: ", stats$total_walkers, "\n",
      "  Completed: ", stats$completed_walkers, "\n",
      "\nSteps:\n",
      "  Total: ", stats$total_steps, "\n",
      "  Min: ", stats$min_steps, "\n",
      "  Max: ", stats$max_steps, "\n",
      "  Mean: ", sprintf("%.1f", stats$mean_steps), "\n",
      "  Median: ", stats$median_steps, "\n",
      "\nPerformance:\n",
      "  Elapsed Time: ", sprintf("%.2f", stats$elapsed_time_secs), "s\n",
      "\nTermination Reasons:\n",
      paste(sprintf("  %s: %d", names(stats$termination_reasons),
                    stats$termination_reasons), collapse = "\n")
    )
  })

  # Backend information with dynamic simulation data
  output$backend_info <- renderText({
    is_webr <- exists(".webr_env", envir = .GlobalEnv) ||
               identical(Sys.getenv("WEBR"), "1")

    # Static environment info
    env_info <- paste0(
      "=== ENVIRONMENT ===\n",
      "Runtime: WebR/Browser\n",
      "Mode: sync sequential (browser limitation)\n",
      "\n=== PACKAGE VERSIONS ===\n",
      "randomwalk: ", packageVersion("randomwalk"), "\n",
      if (requireNamespace("mirai", quietly = TRUE)) {
        paste0("mirai: ", packageVersion("mirai"), "\n")
      } else "",
      if (requireNamespace("nanonext", quietly = TRUE)) {
        paste0("nanonext: ", packageVersion("nanonext"), "\n")
      } else "",
      if (requireNamespace("crew", quietly = TRUE)) {
        paste0("crew: ", packageVersion("crew"), "\n")
      } else ""
    )

    # Dynamic simulation data (if available)
    result <- sim_result()
    if (is.null(result)) {
      paste0(
        env_info,
        "\n=== SIMULATION DATA ===\n",
        "No simulation run yet - click 'Run Simulation' to generate data\n",
        "\n=== STATUS ===\n",
        "State: ", sim_state(), "\n",
        "Total Runs: ", sim_count()
      )
    } else {
      stats <- result$statistics

      paste0(
        env_info,
        "\n=== SIMULATION DATA (Run #", sim_count(), ") ===\n",
        "State: ", sim_state(), "\n",
        if (!is.null(sim_start_time()) && !is.null(sim_end_time())) {
          paste0(
            "Started: ", format(sim_start_time(), "%H:%M:%S"), "\n",
            "Finished: ", format(sim_end_time(), "%H:%M:%S"), "\n",
            "Elapsed: ", sprintf("%.2f", as.numeric(difftime(sim_end_time(), sim_start_time(), units = "secs"))), " seconds\n"
          )
        } else "",
        "\n=== GRID STATE ===\n",
        "Size: ", stats$grid_size, "×", stats$grid_size, "\n",
        "Black Pixels: ", stats$black_pixels, " (", sprintf("%.1f%%", stats$black_percentage), ")\n",
        "White Pixels: ", stats$grid_size^2 - stats$black_pixels, " (", sprintf("%.1f%%", 100 - stats$black_percentage), ")\n",
        "\n=== WALKER STATISTICS ===\n",
        "Total: ", stats$total_walkers, "\n",
        "Completed: ", stats$completed_walkers, "\n",
        "Steps (Total): ", stats$total_steps, "\n",
        "Steps (Mean): ", sprintf("%.1f", stats$mean_steps), "\n",
        "Steps (Median): ", stats$median_steps, "\n",
        "Steps (Min): ", stats$min_steps, "\n",
        "Steps (Max): ", stats$max_steps, "\n",
        "\n=== TERMINATION REASONS ===\n",
        if (!is.null(stats$termination_reasons) && length(stats$termination_reasons) > 0) {
          paste(sprintf("%s: %d (%.1f%%)",
                        names(stats$termination_reasons),
                        stats$termination_reasons,
                        100 * stats$termination_reasons / stats$total_walkers),
                collapse = "\n")
        } else {
          "No data"
        },
        "\n=== PERFORMANCE ===\n",
        "Elapsed: ", sprintf("%.2f", stats$elapsed_time_secs), " seconds\n",
        "Steps/Second: ", sprintf("%.0f", stats$total_steps / max(stats$elapsed_time_secs, 0.001)), "\n"
      )
    }
  })

  # Package details
  output$package_details <- renderText({
    paste0(
      "=== PACKAGE SOURCES ===\n\n",
      "randomwalk:\n",
      "  https://johngavin.github.io/randomwalk/\n\n",
      "WebR packages (NOT installed - parallel processing disabled):\n",
      "  nanonext & mirai: https://r-lib.r-universe.dev\n",
      "  Note: Not compatible with WebR - causes browser version to run sync sequential\n\n",
      "=== PROCESSING MODES ===\n\n",
      "THIS BROWSER VERSION (WebR):\n",
      "  - Sync Sequential Mode (workers=0)\n",
      "  - Processes one walker at a time sequentially\n",
      "  - No parallel processing available\n",
      "  - All computation happens client-side in browser\n\n",
      "R CODE OUTSIDE WEBPAGE (Native R):\n",
      "  - Async Parallel Mode (workers=2+)\n",
      "  - Uses crew backend for async parallelization\n",
      "  - crew internally uses mirai for parallel processing\n",
      "  - Can process multiple walkers simultaneously\n",
      "  - Full async/parallel capabilities when running randomwalk package\n",
      "    in regular R (not in browser)\n\n",
      "=== SYNC MODES (Both Environments) ===\n\n",
      "sync_mode='static' (current):\n",
      "  - Workers get grid snapshots\n",
      "  - Used in native R async parallel mode\n",
      "  - Not applicable in browser (no workers)\n\n",
      "sync_mode='dynamic' (future):\n",
      "  - Real-time broadcasting via nanonext\n",
      "  - Issue #130 - future work\n",
      "  - Would require nanonext WebR compatibility"
    )
  })
}

# ============================================================================
# Run Shiny App
# ============================================================================

shinyApp(ui = ui, server = server)
```

## About This Dashboard

This dashboard provides an **interactive Shiny application** for random walk simulations running in your browser using WebAssembly (WebR).

> **Browser Mode:** This dashboard runs in **sync sequential mode** only (workers=0). Each walker is processed one at a time. This is a WebR/browser limitation.

## Features

### Visualization Types
- **Grid Display**: Black pixel patterns showing the final grid state
- **Walker Paths**: Trajectories with selectable first N and last N paths
- **Distributions**: Path length histograms overall and by termination reason
- **Statistics**: Simulation metrics and timing
- **Debug Info**: Package versions and environment detection

### Simulation Controls
- **Grid Size**: 20-400 pixels
- **Number of Walkers**: 1-5000 walkers
- **Max Steps**: 100-10000 steps per walker
- **Neighborhood**: 4-hood (cardinal) or 8-hood (diagonal)
- **Boundary**: Terminate at edge or wrap (toroidal)

## Dashboard Comparison

Both browser dashboards run in **sync sequential mode** (identical visualization capabilities):

| Feature | [Comprehensive Dashboard](dashboard_comprehensive.html){target="_blank"} | This Dashboard |
|---------|-----------------------------------|---------------|
| Default walkers | 2000 | 2000 |
| Output tabs | 6 (Fractal, Paths, Distributions, Stats, Debug, Notes) | 6 (Grid, Paths, Distributions, Stats, Details, Debug) |
| Statistics | Comprehensive | Comprehensive |
| Best for | Full exploration | Quick demos |

## Try It Yourself

Experiment with different parameters:

- **Larger grid**: More exploration space (try 200-400)
- **Fewer walkers**: Faster simulation (try 100-500)
- **8-hood**: Enable diagonal movement for different patterns
- **Wrap boundary**: Toroidal grid (no edge termination)

## Related Links

- [randomwalk Package Documentation](https://johngavin.github.io/randomwalk/)
- [Comprehensive Dashboard](dashboard_comprehensive.html)

## Session Info

```{r session-info}
sessionInfo()
```

## Version Information

```{r version-info}
#| echo: false
#| results: asis

# Git commit information (captured at build time)
git_sha <- tryCatch(
  system("git rev-parse --short HEAD 2>/dev/null", intern = TRUE),
  error = function(e) "N/A"
)
git_sha <- if (length(git_sha) == 0 || git_sha == "") "N/A" else git_sha

# Package version
pkg_ver <- tryCatch(
  as.character(packageVersion("randomwalk")),
  error = function(e) "N/A"
)

# Build timestamp
build_ts <- format(Sys.time(), "%Y-%m-%d %H:%M %Z")

# Output as markdown table
cat("| Item | Value |\n")
cat("|------|-------|\n")
cat(sprintf("| **Git SHA** | `%s` |\n", git_sha))
cat(sprintf("| **Package** | randomwalk %s |\n", pkg_ver))
cat(sprintf("| **R Version** | %s |\n", R.version.string))
cat(sprintf("| **Built** | %s |\n", build_ts))
```
