---
title: "Simulation Profiling and Performance Analysis"
description: "Profiling results identifying performance bottlenecks in random walk simulation"
author: "randomwalk package"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
    embed-resources: false
execute:
  eval: true
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation Profiling and Performance}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
 echo = TRUE,
 message = FALSE,
 warning = FALSE,
 fig.width = 10,
 fig.height = 6
)

library(dplyr)
library(ggplot2)
library(DT)
```

## Overview
This vignette documents profiling results for the randomwalk simulation,
identifying performance bottlenecks and providing optimization recommendations.

## Profiling Methodology

We profiled the synchronous simulation mode using `profvis` with:

- **Small grid**: 30×30 pixels, 50 walkers
- **Medium grid**: 50×50 pixels, 100 walkers
- **Sampling interval**: 5ms

```r
# Example profiling code
library(profvis)
library(randomwalk)

prof <- profvis({
  result <- run_simulation(
    grid_size = 50,
    n_walkers = 100,
    workers = 0,
    verbose = FALSE
  )
}, interval = 0.005)
```

## Top Functions by Execution Time

Based on profiling the medium grid configuration (50×50, 100 walkers):
```{r top-functions}
top_functions <- tibble::tribble(
 ~Rank, ~Function, ~Time_ms, ~Pct_Total, ~Context,
 1, "run_simulation", 17063, 23.1, "Main simulation orchestration",
 2, "step_walker", 10839, 14.7, "Individual walker movement logic",
 3, "match.arg", 6033, 8.2, "R argument validation overhead",
 4, "check_termination_fast", 4340, 5.9, "Walker termination checks",
 5, "logger::log_trace", 3215, 4.4, "Logging overhead"
)

DT::datatable(
 top_functions,
 caption = "Top 5 functions by execution time (50×50 grid, 100 walkers)",
 options = list(pageLength = 10, dom = 'tip'),
 rownames = FALSE
)
```

## Overhead Analysis

```{r overhead-analysis}
overhead <- tibble::tribble(
 ~Category, ~Pct_Time, ~Time_ms, ~Optimization,
 "Core simulation logic", 22.9, 16928, "Inherent - optimize algorithms",
 "Logging", 11.2, 8261, "Use verbose=FALSE in production",
 "Argument validation (match.arg)", 8.2, 6033, "Pre-validate inputs",
 "Termination checks", 5.9, 4340, "Already optimized (vectorized)",
 "Neighbor operations", 7.0, 5170, "Consider caching neighbor lists"
)

ggplot(overhead, aes(x = reorder(Category, Pct_Time), y = Pct_Time, fill = Category)) +
 geom_col() +
 geom_text(aes(label = paste0(Pct_Time, "%")), hjust = -0.1, size = 3.5) +
 coord_flip() +
 scale_y_continuous(limits = c(0, 30)) +
 labs(
   title = "Execution Time Breakdown by Category",
   subtitle = "50×50 grid, 100 walkers",
   x = NULL,
   y = "Percentage of Total Time"
 ) +
 theme_minimal() +
 theme(legend.position = "none")
```

## Benchmarks: Scaling Analysis

### Grid Size and Walker Count Impact
```{r benchmark-table}
benchmarks <- tibble::tribble(
 ~Grid_Size, ~Walkers, ~Median_ms, ~Min_ms, ~Throughput,
 "20×20", 20, 78.8, 47.4, "12.1/sec",
 "30×30", 20, 83.5, 78.8, "6.2/sec",
 "50×50", 20, 313.6, 204.6, "3.3/sec",
 "20×20", 50, 76.1, 64.1, "13.1/sec",
 "30×30", 50, 173.9, 157.3, "5.5/sec",
 "50×50", 50, 500.3, 395.7, "2.0/sec",
 "20×20", 100, 128.8, 77.3, "8.7/sec",
 "30×30", 100, 277.9, 277.5, "3.5/sec",
 "50×50", 100, 1189.1, 991.7, "0.9/sec"
)

DT::datatable(
 benchmarks,
 caption = "Benchmark results across grid sizes and walker counts",
 options = list(pageLength = 10, dom = 'tip'),
 rownames = FALSE
)
```

### Scaling Visualization
```{r scaling-plot}
bench_data <- tibble::tribble(
 ~grid_size, ~walkers, ~time_ms,
 20, 20, 78.8,
 30, 20, 83.5,
 50, 20, 313.6,
 20, 50, 76.1,
 30, 50, 173.9,
 50, 50, 500.3,
 20, 100, 128.8,
 30, 100, 277.9,
 50, 100, 1189.1
)

ggplot(bench_data, aes(x = grid_size, y = time_ms, color = factor(walkers), group = walkers)) +
 geom_line(linewidth = 1) +
 geom_point(size = 3) +
 scale_y_log10() +
 labs(
   title = "Simulation Time vs Grid Size",
   subtitle = "Log scale - lines show different walker counts",
   x = "Grid Size (pixels)",
   y = "Median Time (ms, log scale)",
   color = "Walkers"
 ) +
 theme_minimal()
```

### Complexity Analysis

- **Grid size scaling**: O(n^1.8) where n = grid dimension
- **Walker count scaling**: O(w^0.7) where w = number of walkers

The sub-linear scaling with walker count (0.7 exponent) indicates efficient
handling of multiple walkers, likely due to early termination when walkers
hit existing black pixels.

## Sync vs Async Comparison

**Critical finding**: Async mode is significantly slower for typical workloads.
```{r sync-async}
async_compare <- tibble::tribble(
 ~Mode, ~Median_ms, ~Speedup,
 "Sync (workers=0)", 173.5, "1.0x (baseline)",
 "Async (workers=2)", 2229.6, "0.08x (12.85x slower)"
)

DT::datatable(
 async_compare,
 caption = "Sync vs Async performance (30×30 grid, 50 walkers)",
 options = list(dom = 't'),
 rownames = FALSE
)
```

### Why Async is Slower

For typical workloads (grids ≤50×50, walkers ≤100):

1. **Parallelization overhead dominates** - Worker spawning, task distribution,
   and result collection take longer than the actual computation
2. **Memory pressure** - Each worker needs its own grid copy
3. **Small problem size** - Not enough work to amortize coordination costs

### When to Use Each Mode

| Scenario | Recommended Mode | Reason |
|----------|-----------------|--------|
| Interactive dashboards | Sync (`workers=0`) | Sub-second response |
| Grids ≤100×100 | Sync | Overhead not worth it |
| Grids >100×100 | Test both | May benefit from async |
| Batch processing | Sync per-sim, parallel over configs | External parallelism |

## Optimization Recommendations

### High Impact (>5% improvement)

1. **Disable logging in production** (~11% savings)
   ```r
   run_simulation(..., verbose = FALSE)
   ```

2. **Reduce match.arg overhead** (~8% savings)
   - Pre-validate inputs before hot loops
   - Consider `.Internal()` for critical paths

### Medium Impact (2-5% improvement)

3. **Cache neighbor lists** (~3-7% savings)
   - Pre-compute neighbors for each grid cell
   - Avoid repeated `get_neighbors()` calls

4. **Vectorize termination checks** (~2-4% savings)
   - Batch checks across multiple walkers

### Low Impact (<2% improvement)

5. **Optimize grid indexing**
   - Already uses efficient matrix operations
   - Further optimization yields diminishing returns

## Interactive Profiling

The raw profvis data is saved in the package for interactive exploration:

```r
# Load and view interactive flame graph
prof <- readRDS(system.file("extdata", "prof_sync_medium.rds", package = "randomwalk"))
print(prof)  # Opens interactive viewer
```

## Session Info

```{r session-info}
sessionInfo()
```
