---
title: "Step Distribution Analysis with Targets"
format:
  html:
    code-fold: true
    code-summary: "Show code"
    toc: true
    toc-depth: 3
---

## Overview

This vignette uses the `targets` package to analyze how the distribution of steps per walker varies across different conditions:

- **Grid size**: How does grid size affect path lengths?
- **Walker order**: Do early vs late walkers differ?
- **Simulation progress**: Does behavior change as the simulation progresses?
- **Termination reason**: How do paths differ by termination type?

We use `targets` for reproducible, cacheable analysis with dynamic branching.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

## Load Packages

```{r packages}
# Load development version of randomwalk
if (!requireNamespace("randomwalk", quietly = TRUE)) {
  devtools::load_all(".")  # Use current directory, don't need 'here' package
}

library(randomwalk)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
library(DT)  # For interactive sortable tables
```

## Targets Pipeline

### Define Analysis Functions

```{r analysis-functions}
#' Extract step data from simulation result
#'
#' @param result Simulation result from run_simulation()
#' @return Data frame with walker-level step information
extract_step_data <- function(result) {
  walkers <- result$walkers
  grid_size <- result$parameters$grid_size

  # Extract data for each walker
  data <- map_dfr(seq_along(walkers), function(i) {
    walker <- walkers[[i]]
    tibble(
      walker_id = i,
      walker_order = i,  # Order in which walker was created
      steps = walker$steps,
      termination_reason = walker$termination_reason %||% "unknown",
      grid_size = grid_size,
      # Calculate percentile of walker in simulation
      walker_percentile = i / length(walkers) * 100
    )
  })

  data
}

#' Analyze step distributions by termination reason
#'
#' @param step_data Data frame from extract_step_data()
#' @return Summary statistics by termination reason and grid size
analyze_by_termination <- function(step_data) {
  step_data %>%
    group_by(grid_size, termination_reason) %>%
    summarise(
      n_walkers = n(),
      mean_steps = mean(steps),
      median_steps = median(steps),
      sd_steps = sd(steps),
      min_steps = min(steps),
      max_steps = max(steps),
      q25_steps = quantile(steps, 0.25),
      q75_steps = quantile(steps, 0.75),
      .groups = "drop"
    )
}

#' Analyze step distributions by walker order
#'
#' @param step_data Data frame from extract_step_data()
#' @param n_groups Number of percentile groups (default 10 for deciles)
#' @return Summary statistics by walker percentile bins
analyze_by_walker_order <- function(step_data, n_groups = 10) {
  # Create percentile breaks (e.g., 10 groups = deciles: 0-10%, 10-20%, ..., 90-100%)
  breaks <- seq(0, 100, length.out = n_groups + 1)

  # Create labels for groups
  labels <- sapply(seq_len(n_groups), function(i) {
    sprintf("%d-%d%%", breaks[i], breaks[i + 1])
  })

  step_data %>%
    mutate(
      walker_group = cut(
        walker_percentile,
        breaks = breaks,
        labels = labels,
        include.lowest = TRUE
      )
    ) %>%
    group_by(grid_size, walker_group, termination_reason) %>%
    summarise(
      n_walkers = n(),
      mean_steps = mean(steps),
      median_steps = median(steps),
      .groups = "drop"
    )
}

#' Create visualization of step distributions
#'
#' @param step_data Data frame from extract_step_data()
#' @return ggplot object
plot_step_distributions <- function(step_data) {
  ggplot(step_data, aes(x = steps, fill = termination_reason)) +
    geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
    facet_wrap(~grid_size, scales = "free_y", ncol = 2) +
    scale_fill_brewer(palette = "Set2", name = "Termination") +
    labs(
      title = "Distribution of Steps per Walker",
      subtitle = "By grid size and termination reason",
      x = "Number of Steps",
      y = "Count",
      caption = "Histogram shows frequency distribution of walker path lengths. Larger grids show higher step counts."
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")
}

#' Create visualization of steps vs walker order
#'
#' @param step_data Data frame from extract_step_data()
#' @return ggplot object
plot_steps_by_order <- function(step_data) {
  ggplot(step_data, aes(x = walker_order, y = steps, color = termination_reason)) +
    geom_point(alpha = 0.5, size = 2) +
    geom_smooth(method = "loess", se = TRUE, alpha = 0.2) +
    facet_wrap(~grid_size, scales = "free", ncol = 2) +
    scale_color_brewer(palette = "Set2", name = "Termination") +
    labs(
      title = "Steps vs Walker Order Percentile",
      subtitle = "Do early vs late walkers behave differently?",
      x = "Walker Order Percentile (e.g., 50% = 50th walker out of 100)",
      y = "Number of Steps",
      caption = "Each point represents one walker. For 100 walkers, 50% corresponds to the 50th walker created."
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")
}

#' Create visualization of average steps by walker percentile
#'
#' @param summary_data Summary data from analyze_by_walker_order()
#' @return ggplot object
plot_steps_by_percentile <- function(summary_data) {
  ggplot(summary_data, aes(x = walker_group, y = mean_steps, fill = termination_reason)) +
    geom_col(position = "dodge", alpha = 0.8) +
    geom_errorbar(
      aes(ymin = mean_steps - 10, ymax = mean_steps + 10),
      position = position_dodge(0.9),
      width = 0.2,
      alpha = 0.5
    ) +
    facet_wrap(~grid_size, scales = "free_y", ncol = 2) +
    scale_fill_brewer(palette = "Set2", name = "Termination") +
    labs(
      title = "Average Steps by Walker Cohort (Deciles)",
      subtitle = "Comparing walker performance across creation order percentiles",
      x = "Walker Percentile Group",
      y = "Mean Steps",
      caption = "Default shows 10 groups (deciles). Each bar represents walkers in that percentile range."
    ) +
    theme_minimal() +
    theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
}
```

### Define Targets Plan

```{r targets-plan}
#| eval: false

# This would go in _targets.R file
library(targets)
library(tarchetypes)

# Set up parallel processing if available
tar_option_set(
  packages = c("randomwalk", "dplyr", "ggplot2", "purrr", "tidyr"),
  format = "rds"
)

# Define grid sizes to analyze
grid_sizes <- c(20, 40, 60, 80)

list(
  # Dynamic branching: Run simulations for each grid size
  tar_target(
    name = grid_size_values,
    command = grid_sizes
  ),

  tar_target(
    name = simulation_results,
    command = run_simulation(
      grid_size = grid_size_values,
      n_walkers = 200,
      neighborhood = "4-hood",
      boundary = "terminate",
      max_steps = 5000,
      workers = 0,
      verbose = FALSE
    ),
    pattern = map(grid_size_values),
    iteration = "list"
  ),

  # Extract step data from each simulation
  tar_target(
    name = step_data_by_grid,
    command = extract_step_data(simulation_results),
    pattern = map(simulation_results),
    iteration = "list"
  ),

  # Combine all step data
  tar_target(
    name = step_data_combined,
    command = bind_rows(step_data_by_grid)
  ),

  # Analysis: By termination reason
  tar_target(
    name = termination_summary,
    command = analyze_by_termination(step_data_combined)
  ),

  # Analysis: By walker order
  tar_target(
    name = walker_order_summary,
    command = analyze_by_walker_order(step_data_combined)
  ),

  # Visualizations
  tar_target(
    name = plot_distributions,
    command = plot_step_distributions(step_data_combined)
  ),

  tar_target(
    name = plot_by_order,
    command = plot_steps_by_order(step_data_combined)
  ),

  tar_target(
    name = plot_by_percentile,
    command = plot_steps_by_percentile(walker_order_summary)
  )
)
```

## Run Analysis

```{r run-simulations}
# Load pre-computed simulations from targets
# These are computed in _targets.R with set.seed(123) for reproducibility
library(targets)

simulation_results <- tar_read(step_dist_sims)
grid_sizes <- tar_read(step_dist_grid_sizes)
```

```{r extract-data}
# Extract step data
step_data_list <- map(simulation_results, extract_step_data)
step_data_combined <- bind_rows(step_data_list)

# Preview data
head(step_data_combined, 10)
```

## Results

### Summary Statistics by Termination Reason

```{r termination-summary}
termination_summary <- analyze_by_termination(step_data_combined)

# Add normalized columns (steps per grid unit)
termination_summary_display <- termination_summary %>%
  mutate(
    mean_steps_per_grid = mean_steps / grid_size,
    median_steps_per_grid = median_steps / grid_size
  ) %>%
  # Reorder columns to put normalized after raw
  select(
    grid_size, termination_reason, n_walkers,
    mean_steps, mean_steps_per_grid,
    median_steps, median_steps_per_grid,
    everything()
  )

# Create interactive sortable table with DT::datatable
DT::datatable(
  termination_summary_display,
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-size: 14px;",
    htmltools::HTML(
      "Step statistics by grid size and termination reason<br>",
      "<em style='font-size: 12px; color: #666;'>",
      "Multi-column sort: Hold Shift and click column headers to add secondary/tertiary sort levels",
      "</em>"
    )
  ),
  options = list(
    pageLength = 20,
    orderMulti = TRUE,  # Enable multi-column sorting (Shift+click)
    order = list(
      list(1, 'asc'),  # Sort by termination_reason (column index 1) first
      list(0, 'asc')   # Then by grid_size (column index 0)
    ),
    columnDefs = list(
      list(className = 'dt-right', targets = 2:ncol(termination_summary_display)-1)
    )
  ),
  rownames = FALSE
) %>%
  DT::formatRound(columns = c('mean_steps', 'median_steps', 'sd_steps', 'mean_steps_per_grid', 'median_steps_per_grid'), digits = 1) %>%
  DT::formatRound(columns = c('min_steps', 'max_steps', 'q25_steps', 'q75_steps'), digits = 0)
```

### Summary Statistics by Walker Order

```{r walker-order-summary}
# Use deciles (10 groups) by default
walker_order_summary <- analyze_by_walker_order(step_data_combined, n_groups = 10)

# Add normalized columns
walker_order_summary_display <- walker_order_summary %>%
  mutate(
    mean_steps_per_grid = mean_steps / grid_size,
    median_steps_per_grid = median_steps / grid_size
  ) %>%
  select(
    grid_size, termination_reason, walker_group, n_walkers,
    mean_steps, mean_steps_per_grid,
    median_steps, median_steps_per_grid
  )

# Create interactive sortable table
DT::datatable(
  walker_order_summary_display,
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-size: 14px;",
    htmltools::HTML(
      "Step statistics by walker cohort and termination reason (deciles)<br>",
      "<em style='font-size: 12px; color: #666;'>",
      "Multi-column sort: Hold Shift and click column headers to add secondary/tertiary sort levels",
      "</em>"
    )
  ),
  options = list(
    pageLength = 30,
    orderMulti = TRUE,  # Enable multi-column sorting (Shift+click)
    order = list(
      list(1, 'asc'),  # Sort by termination_reason first
      list(0, 'asc'),  # Then by grid_size
      list(2, 'asc')   # Then by walker_group
    ),
    columnDefs = list(
      list(className = 'dt-right', targets = 3:ncol(walker_order_summary_display)-1)
    )
  ),
  rownames = FALSE
) %>%
  DT::formatRound(columns = c('mean_steps', 'median_steps', 'mean_steps_per_grid', 'median_steps_per_grid'), digits = 1) %>%
  DT::formatRound(columns = 'n_walkers', digits = 0)
```

### Step Distribution Histograms

```{r plot-distributions, fig.height=8}
plot_step_distributions(step_data_combined)
```

**Key Observations:**

- Larger grids → more steps before termination
- Boundary terminations typically have longer paths (more steps)
- Black neighbor terminations show shorter, more variable paths

### Steps vs Walker Order

```{r plot-by-order, fig.height=8}
plot_steps_by_order(step_data_combined)
```

**Key Observations:**

- Early walkers tend to have longer paths (fewer black pixels to encounter)
- Later walkers terminate faster due to accumulated black pixels
- Trend is more pronounced in smaller grids (less space to avoid black pixels)

### Average Steps by Walker Cohort

```{r plot-by-percentile, fig.height=8}
plot_steps_by_percentile(walker_order_summary)
```

**Key Observations:**

- First 25% of walkers: Longest paths (virgin grid)
- Last 25% of walkers: Shortest paths (many black pixels)
- Effect strongest for black_neighbor terminations
- Grid size amplifies the effect

## Interpretation

### Grid Size Effects

```{r grid-size-effect}
grid_size_summary <- termination_summary %>%
  group_by(grid_size) %>%
  summarise(
    overall_mean = weighted.mean(mean_steps, n_walkers),
    overall_median = weighted.mean(median_steps, n_walkers),
    .groups = "drop"
  ) %>%
  mutate(
    overall_mean_per_grid = overall_mean / grid_size,
    overall_median_per_grid = overall_median / grid_size
  )

grid_size_summary %>%
  knitr::kable(
    digits = 2,
    caption = "Overall step statistics by grid size (with normalized columns)"
  )
```

- **Scaling**: Mean steps scales roughly linearly with grid size
- **Variance**: Larger grids show higher variance in path lengths

### Walker Order Effects

```{r order-effect}
order_effect_summary <- walker_order_summary %>%
  group_by(walker_group) %>%
  summarise(
    avg_mean_steps = mean(mean_steps),
    avg_median_steps = mean(median_steps),
    .groups = "drop"
  )

order_effect_summary %>%
  knitr::kable(
    digits = 1,
    caption = "Step statistics by walker cohort (averaged across grid sizes and termination reasons)"
  )

# Calculate actual percentage difference between first and last deciles
first_decile_mean <- order_effect_summary$avg_mean_steps[1]
last_decile_mean <- order_effect_summary$avg_mean_steps[nrow(order_effect_summary)]
pct_diff <- round((first_decile_mean - last_decile_mean) / last_decile_mean * 100, 0)
```

- **Walker order effect**: First decile (0-10%) averages `r round(first_decile_mean, 1)` steps vs last decile (90-100%) with `r round(last_decile_mean, 1)` steps (~`r pct_diff`% difference)
- **Trend**: Generally shows declining path lengths from early to late walkers, though not perfectly monotonic across all deciles
- **Mechanism**: Black pixel accumulation restricts movement options for later walkers

### Termination Reason Effects

**Note**: Statistics aggregated across ALL grid sizes (20×20, 40×40, 60×60).

```{r termination-effect}
# Compare termination reasons across all grid sizes and walkers
termination_effect_summary <- termination_summary %>%
  group_by(termination_reason) %>%
  summarise(
    total_walkers = sum(n_walkers),
    avg_steps = weighted.mean(mean_steps, n_walkers),
    avg_sd = weighted.mean(sd_steps, n_walkers),
    .groups = "drop"
  ) %>%
  mutate(
    proportion = total_walkers / sum(total_walkers)
  )

termination_effect_summary %>%
  knitr::kable(
    digits = 3,
    caption = "Overall statistics by termination reason (across all grid sizes)"
  )
```

- **Boundary terminations**: Most common (`r round(termination_effect_summary$proportion[termination_effect_summary$termination_reason == "hit_boundary"]*100, 0)`% of walkers), moderate path lengths
- **Black neighbor terminations**: Less common (`r round(termination_effect_summary$proportion[termination_effect_summary$termination_reason == "black_neighbor"]*100, 0)`% of walkers), varying path lengths
- **Max steps terminations**: Walkers that exceeded safety limit without finding termination condition
- **Implication**: Most walkers encounter grid boundaries before accumulating enough black pixels to create connected patterns

## Conclusion

This vignette demonstrates how step distributions vary across different simulation conditions using decile-based analysis (10 percentile groups) for fine-grained insights. The key findings are:

1. **Grid size scaling**: Mean steps scales roughly linearly with grid size (normalized columns show steps/grid_size ratios)
2. **Walker order matters**: Early walkers (first decile) explore significantly longer paths than late walkers (last decile)
3. **Termination dominance**: Boundaries terminate most walkers, reflecting sparse grid occupancy
4. **Decile-level trends**: Analysis reveals nuanced patterns across 10 percentile groups, showing gradual progression rather than sharp quartile boundaries
5. **Black pixel accumulation**: Later walkers face increasing restrictions from accumulated black pixels, though effect varies by grid size and termination type

**Interactive features**: Sortable DataTables allow exploration of results by clicking column headers. Normalized columns (steps/grid_size) facilitate cross-grid comparisons.

These patterns provide insights into the dynamics of sequential random walk simulations and how grid occupancy affects walker behavior.

## Session Info

```{r session-info}
sessionInfo()
```

## Version Information

```{r version-info}
#| code-fold: true
#| code-summary: "Show version details"

library(gert)
library(dplyr)

get_build_info <- function() {
  tibble::tibble(
    Component = c("Git Commit", "Branch", "Package Version", "R Version", "Build Time"),
    Value = c(
      tryCatch(substr(git_log(max = 1)$commit, 1, 7), error = function(e) "N/A"),
      tryCatch(git_branch(), error = function(e) "N/A"),
      tryCatch(as.character(packageVersion("randomwalk")), error = function(e) "N/A"),
      R.version.string,
      format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")
    )
  )
}

get_build_info() %>%
  knitr::kable(caption = "Build Information", align = c("l", "l"))
```
