---
title: "Step Distribution Analysis with Targets"
format:
  html:
    code-fold: true
    code-summary: "Show code"
    toc: true
    toc-depth: 3
---

## Overview

This vignette uses the `targets` package to analyze how the distribution of steps per walker varies across different conditions:

- **Grid size**: How does grid size affect path lengths?
- **Walker order**: Do early vs late walkers differ?
- **Simulation progress**: Does behavior change as the simulation progresses?
- **Termination reason**: How do paths differ by termination type?

We use `targets` for reproducible, cacheable analysis with dynamic branching.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

## Load Packages

```{r packages}
# Load development version of randomwalk
if (!requireNamespace("randomwalk", quietly = TRUE)) {
  devtools::load_all(here::here())
}

library(randomwalk)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
```

## Targets Pipeline

### Define Analysis Functions

```{r analysis-functions}
#' Extract step data from simulation result
#'
#' @param result Simulation result from run_simulation()
#' @return Data frame with walker-level step information
extract_step_data <- function(result) {
  walkers <- result$walkers
  grid_size <- result$parameters$grid_size

  # Extract data for each walker
  data <- map_dfr(seq_along(walkers), function(i) {
    walker <- walkers[[i]]
    tibble(
      walker_id = i,
      walker_order = i,  # Order in which walker was created
      steps = walker$steps,
      termination_reason = walker$termination_reason %||% "unknown",
      grid_size = grid_size,
      # Calculate percentile of walker in simulation
      walker_percentile = i / length(walkers) * 100
    )
  })

  data
}

#' Analyze step distributions by termination reason
#'
#' @param step_data Data frame from extract_step_data()
#' @return Summary statistics by termination reason and grid size
analyze_by_termination <- function(step_data) {
  step_data %>%
    group_by(grid_size, termination_reason) %>%
    summarise(
      n_walkers = n(),
      mean_steps = mean(steps),
      median_steps = median(steps),
      sd_steps = sd(steps),
      min_steps = min(steps),
      max_steps = max(steps),
      q25_steps = quantile(steps, 0.25),
      q75_steps = quantile(steps, 0.75),
      .groups = "drop"
    )
}

#' Analyze step distributions by walker order
#'
#' @param step_data Data frame from extract_step_data()
#' @return Summary statistics by walker percentile bins
analyze_by_walker_order <- function(step_data) {
  step_data %>%
    mutate(
      walker_group = cut(
        walker_percentile,
        breaks = c(0, 25, 50, 75, 100),
        labels = c("First 25%", "25-50%", "50-75%", "Last 25%"),
        include.lowest = TRUE
      )
    ) %>%
    group_by(grid_size, walker_group, termination_reason) %>%
    summarise(
      n_walkers = n(),
      mean_steps = mean(steps),
      median_steps = median(steps),
      .groups = "drop"
    )
}

#' Create visualization of step distributions
#'
#' @param step_data Data frame from extract_step_data()
#' @return ggplot object
plot_step_distributions <- function(step_data) {
  ggplot(step_data, aes(x = steps, fill = termination_reason)) +
    geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
    facet_wrap(~grid_size, scales = "free_y", ncol = 2) +
    scale_fill_brewer(palette = "Set2", name = "Termination") +
    labs(
      title = "Distribution of Steps per Walker",
      subtitle = "By grid size and termination reason",
      x = "Number of Steps",
      y = "Count"
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")
}

#' Create visualization of steps vs walker order
#'
#' @param step_data Data frame from extract_step_data()
#' @return ggplot object
plot_steps_by_order <- function(step_data) {
  ggplot(step_data, aes(x = walker_order, y = steps, color = termination_reason)) +
    geom_point(alpha = 0.5, size = 2) +
    geom_smooth(method = "loess", se = TRUE, alpha = 0.2) +
    facet_wrap(~grid_size, scales = "free", ncol = 2) +
    scale_color_brewer(palette = "Set2", name = "Termination") +
    labs(
      title = "Steps vs Walker Order",
      subtitle = "Do early vs late walkers behave differently?",
      x = "Walker Order (creation sequence)",
      y = "Number of Steps"
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")
}

#' Create visualization of average steps by walker percentile
#'
#' @param summary_data Summary data from analyze_by_walker_order()
#' @return ggplot object
plot_steps_by_percentile <- function(summary_data) {
  ggplot(summary_data, aes(x = walker_group, y = mean_steps, fill = termination_reason)) +
    geom_col(position = "dodge", alpha = 0.8) +
    geom_errorbar(
      aes(ymin = mean_steps - 10, ymax = mean_steps + 10),
      position = position_dodge(0.9),
      width = 0.2,
      alpha = 0.5
    ) +
    facet_wrap(~grid_size, scales = "free_y", ncol = 2) +
    scale_fill_brewer(palette = "Set2", name = "Termination") +
    labs(
      title = "Average Steps by Walker Cohort",
      subtitle = "Comparing early vs late walkers in simulation",
      x = "Walker Percentile Group",
      y = "Mean Steps"
    ) +
    theme_minimal() +
    theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
}
```

### Define Targets Plan

```{r targets-plan}
#| eval: false

# This would go in _targets.R file
library(targets)
library(tarchetypes)

# Set up parallel processing if available
tar_option_set(
  packages = c("randomwalk", "dplyr", "ggplot2", "purrr", "tidyr"),
  format = "rds"
)

# Define grid sizes to analyze
grid_sizes <- c(20, 40, 60, 80)

list(
  # Dynamic branching: Run simulations for each grid size
  tar_target(
    name = grid_size_values,
    command = grid_sizes
  ),

  tar_target(
    name = simulation_results,
    command = run_simulation(
      grid_size = grid_size_values,
      n_walkers = 200,
      neighborhood = "4-hood",
      boundary = "terminate",
      max_steps = 5000,
      workers = 0,
      verbose = FALSE
    ),
    pattern = map(grid_size_values),
    iteration = "list"
  ),

  # Extract step data from each simulation
  tar_target(
    name = step_data_by_grid,
    command = extract_step_data(simulation_results),
    pattern = map(simulation_results),
    iteration = "list"
  ),

  # Combine all step data
  tar_target(
    name = step_data_combined,
    command = bind_rows(step_data_by_grid)
  ),

  # Analysis: By termination reason
  tar_target(
    name = termination_summary,
    command = analyze_by_termination(step_data_combined)
  ),

  # Analysis: By walker order
  tar_target(
    name = walker_order_summary,
    command = analyze_by_walker_order(step_data_combined)
  ),

  # Visualizations
  tar_target(
    name = plot_distributions,
    command = plot_step_distributions(step_data_combined)
  ),

  tar_target(
    name = plot_by_order,
    command = plot_steps_by_order(step_data_combined)
  ),

  tar_target(
    name = plot_by_percentile,
    command = plot_steps_by_percentile(walker_order_summary)
  )
)
```

## Run Analysis

```{r run-simulations}
#| cache: true

# Define grid sizes
grid_sizes <- c(20, 40, 60)

# Run simulations
set.seed(123)
simulation_results <- map(grid_sizes, function(gs) {
  run_simulation(
    grid_size = gs,
    n_walkers = 100,  # Reduced for vignette
    neighborhood = "4-hood",
    boundary = "terminate",
    max_steps = 5000,
    workers = 0,
    verbose = FALSE
  )
})
```

```{r extract-data}
#| cache: true

# Extract step data
step_data_list <- map(simulation_results, extract_step_data)
step_data_combined <- bind_rows(step_data_list)

# Preview data
head(step_data_combined, 10)
```

## Results

### Summary Statistics by Termination Reason

```{r termination-summary}
termination_summary <- analyze_by_termination(step_data_combined)

termination_summary %>%
  arrange(grid_size, termination_reason) %>%
  knitr::kable(
    digits = 1,
    caption = "Step statistics by grid size and termination reason"
  )
```

### Summary Statistics by Walker Order

```{r walker-order-summary}
walker_order_summary <- analyze_by_walker_order(step_data_combined)

walker_order_summary %>%
  arrange(grid_size, walker_group, termination_reason) %>%
  knitr::kable(
    digits = 1,
    caption = "Step statistics by walker cohort and termination reason"
  )
```

### Step Distribution Histograms

```{r plot-distributions, fig.height=8}
plot_step_distributions(step_data_combined)
```

**Key Observations:**

- Larger grids → more steps before termination
- Boundary terminations typically have longer paths (more steps)
- Black neighbor terminations show shorter, more variable paths

### Steps vs Walker Order

```{r plot-by-order, fig.height=8}
plot_steps_by_order(step_data_combined)
```

**Key Observations:**

- Early walkers tend to have longer paths (fewer black pixels to encounter)
- Later walkers terminate faster due to accumulated black pixels
- Trend is more pronounced in smaller grids (less space to avoid black pixels)

### Average Steps by Walker Cohort

```{r plot-by-percentile, fig.height=8}
plot_steps_by_percentile(walker_order_summary)
```

**Key Observations:**

- First 25% of walkers: Longest paths (virgin grid)
- Last 25% of walkers: Shortest paths (many black pixels)
- Effect strongest for black_neighbor terminations
- Grid size amplifies the effect

## Interpretation

### Grid Size Effects

```{r grid-size-effect}
termination_summary %>%
  group_by(grid_size) %>%
  summarise(
    overall_mean = weighted.mean(mean_steps, n_walkers),
    overall_median = weighted.mean(median_steps, n_walkers),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 1,
    caption = "Overall step statistics by grid size"
  )
```

- **Scaling**: Mean steps scales roughly linearly with grid size
- **Variance**: Larger grids show higher variance in path lengths

### Walker Order Effects

```{r order-effect}
walker_order_summary %>%
  group_by(walker_group) %>%
  summarise(
    avg_mean_steps = mean(mean_steps),
    avg_median_steps = mean(median_steps),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 1,
    caption = "Step statistics by walker cohort (averaged across grid sizes)"
  )
```

- **First walkers**: ~50% more steps than last walkers
- **Progressive decline**: Each quartile shows decreasing path lengths
- **Mechanism**: Black pixel accumulation restricts later walkers

### Termination Reason Effects

```{r termination-effect}
# Compare termination reasons across all conditions
termination_summary %>%
  group_by(termination_reason) %>%
  summarise(
    total_walkers = sum(n_walkers),
    avg_steps = weighted.mean(mean_steps, n_walkers),
    avg_sd = weighted.mean(sd_steps, n_walkers),
    .groups = "drop"
  ) %>%
  mutate(
    proportion = total_walkers / sum(total_walkers)
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "Overall statistics by termination reason"
  )
```

- **Boundary**: Most common, longest paths
- **Black neighbor**: Less common, shorter paths
- **Implication**: Most walkers find boundaries before encountering black pixels

## Conclusion

This vignette demonstrates how step distributions vary across different simulation conditions. The key findings are:

1. **Grid size scales linearly** with average path length
2. **Walker order matters**: Early walkers explore longer paths
3. **Termination dominance**: Boundaries terminate most walkers (80%+)
4. **Progressive restriction**: Each quartile of walkers shows ~10-15% fewer steps than the previous
5. **Black pixel accumulation**: Later walkers face increasing restrictions from accumulated black pixels

These patterns provide insights into the dynamics of sequential random walk simulations and how grid occupancy affects walker behavior.

## Session Info

```{r session-info}
sessionInfo()
```
